---
title: "Jobs -> Payrolls miss but household survey strong again"
date: 2022-01-07T21:13:14-05:00
draft: false
---
&nbsp;
&nbsp;



Job gains in December came in below expectations. A more holistic view of the data, however, shows that the recovery in the labor market remains on track, similar to the dynamic of the November report. Establishment survey metrics continue to look weaker relative to the metrics based on the Household survey. Interestingly, some of this disparity may be coming from non-sampling error due to the establishment survey's inability to capture employment generated by new firms. The BLS tries to make a couple adjustments to address this error, but it's possible that the level of new business creation has been so elevated recently that the BLS' adjustments do not fully account for it. 

```{r setup, include=FALSE}

library(tidyverse)
library(blscrapeR)
library(janitor)
library(kableExtra)
library(formattable)
library(gganimate)
library(quantmod)
library(scales)
library(magick)
library(Cairo)
library(lubridate)

windowsFonts("Open Sans" = windowsFont("Open Sans"))

colors <- c('paleblue' = "#a6cee3", 'blue' = "#1f78b4", 'green' = "#33a02c", 'palegreen' = "#b2df8a",
            'red' = "#e31a1c", 'palered' = "#fb9a99", 'orange' = "#ff7f00", 'paleorange' = "#fdbf6f",
            'purple' = "#6a3d9a", 'palepurple' = "#cab2d6", 'border grey' = "#808080", 'brown' = '#b15928',
            'text' = "#222222" , 'light grey' = "#CDCDCD")

cpal <- function(...) {
  cols <- c(...)
  if (is.null(cols))
    return(colors)
  colors[[cols]]
}

my_theme <- theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.key = element_blank(),
    legend.key.size = unit(0.37, "cm"),
    legend.background = element_blank(),
    legend.margin = margin(t = -0.5, unit = "cm"),
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = cpal("light grey"), size = 0.21),
    panel.border = element_rect(fill = NA, color = cpal("border grey"), size = 0.3),
    plot.title = element_text(size = 21, color = "black", family = "Open Sans Light"),
    plot.subtitle = element_text(size = 14, color = cpal("text"), family = "Open Sans"),
    plot.caption = element_text(size = 12.5, color = cpal("text"), family = "Open Sans"),
    text = element_text(size = 15, color = cpal("text"), family = "Open Sans"),
    legend.text = element_text(size = 14, color = cpal("text"), family = "Open Sans"),
    axis.text.x = element_text(size = 14, color = cpal("text"), family = "Open Sans",
                               margin = unit(c(0.35, 0.25, 0.25, 0.25), "cm")),
    axis.text.y = element_text(size = 14, color = cpal("text"), family = "Open Sans",
                               margin = unit(c(0.35, 0.25, 0.25, 0.25), "cm")),
    axis.title.y = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.length = unit(-0.15, "cm"),
    axis.ticks = element_line(color = cpal("border grey"), size = 0.3),
    plot.margin = unit(c(0.3, 0.5, 0.3, 0.05), "cm")
  )
```


```{r echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE}

ctitle_1 <- "Estimates of monthly change in nonfarm payrolls"
unit_label <- "Payrolls, thousands, seasonally adjusted"
source_1 <- "Source: BLS, @benbakkum."

## https://www.fxstreet.com/macroeconomics/economic-indicator/nfp
## https://www.bls.gov/web/empsit/cesnaicsrev.htm

read_csv("D:\\Documents\\R directory\\labor market\\nfp_revisions.csv") %>% 
  mutate(month = paste0(gsub("\\.", "",Month)," ", Year),
         month = as.yearmon(month, "%b %Y"),
         month = as.Date(month)) %>% 
  select(-Month, -Year) %>% 
  pivot_longer(-month, values_to = "values", names_to = "estimate") %>%  ## edit this
  mutate(date_released = case_when(estimate == "1st" ~ month,
                                   estimate == "2nd" ~ month %m+% months(1),
                                   estimate == "3rd" ~ month %m+% months(2),
                                   estimate == "Consensus" ~ month),
         released = format(date_released, "%b '%y"),
         estimate = ifelse(estimate == "3rd", "Final", estimate),
         index = as.numeric(month) - 18500,
         rank_index = dense_rank(index) - 16, ## need to edit this number
         txt_labels = paste0(values, "k")) %>%
  filter(month > "2021-04-30") %>%
  rename(Estimate = estimate) %>% 
  drop_na() -> df

df$Estimate <- factor(df$Estimate, levels = c("Consensus", "1st", "2nd", "Final"))


df %>% 
  ggplot(aes(rank_index, values, color = Estimate, shape = Estimate)) +
  geom_point(size = 6, alpha = 0.6) +
  geom_text(data = df %>% filter(Estimate == "1st"), aes(2.2, -95, label = released), 
            color = "#808080", size = 13, alpha = .8, 
            family = "Open Sans") +
  geom_text(data = df, aes(rank_index, values, label = txt_labels, color = Estimate),
            size = 6, nudge_x = 0.5, show.legend = F) +
  geom_hline(yintercept = 0, linetype = "dashed", color = cpal("border grey")) +
  scale_color_manual(values = c(cpal("border grey"), cpal("blue"), cpal("orange"), cpal("green"))) +
  scale_shape_manual(values = c(18, 16, 17, 15)) +
  scale_x_continuous(breaks = seq(1,8), 
                     labels = unique(df$released)) +
  scale_y_continuous(breaks = c(-200, 0, 200, 400, 600, 800, 1000),
                     limits = c(-200, 1100)) +
  theme(legend.position = "bottom",
    legend.key = element_blank(),
    legend.key.size = unit(0.37, "cm"),
    legend.background = element_blank(),
    legend.margin = margin(t = -0.2, unit = "cm"),
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = cpal("light grey"), size = 0.21),
    panel.border = element_rect(fill = NA, color = cpal("border grey"), size = 0.3),
    plot.title = element_text(size = 21, color = "black", family = "Open Sans Light"),
    plot.subtitle = element_text(size = 14, color = cpal("text"), family = "Open Sans"),
    plot.caption = element_text(size = 12.5, color = cpal("text"), family = "Open Sans"),
    text = element_text(size = 15, color = cpal("text"), family = "Open Sans"),
    legend.text = element_text(size = 13.5, color = cpal("text"), family = "Open Sans"),
    axis.text.x = element_text(size = 14, color = cpal("text"), family = "Open Sans",
                               margin = unit(c(0.35, 0.25, 0.25, 0.25), "cm")),
    axis.text.y = element_text(size = 14, color = cpal("text"), family = "Open Sans",
                               margin = unit(c(0.35, 0.25, 0.25, 0.25), "cm")),
    axis.title.y = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.length = unit(-0.15, "cm"),
    axis.ticks = element_line(color = cpal("border grey"), size = 0.3),
    plot.margin = unit(c(0.3, 0.5, 0.3, 0.05), "cm")) +
    labs(title = ctitle_1, subtitle = unit_label, caption = source_1) +
    transition_states(date_released, wrap = FALSE) +
    shadow_mark(exclude_layer = c(2, 3)) -> chart1


anim <- animate(chart1, renderer = gifski_renderer(loop = T), nframe = 230,
                 width = 615, height = 420, type = "cairo", end_pause = 90)

anim
  
```


On a non-seasonally adjusted basis, the net number of payrolls added in December was higher than the average of all the Decembers of the 10 years preceding covid. On a seasonally adjusted basis, December job gains were average based on longer-term history, but appear soft in the context of gains earlier in 2021. 

```{r echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE}
dat <- bls_api(c("CEU0000000001", "CES0000000001"), 
              startyear = 2005, 
              endyear = 2021,
              Sys.getenv("BLS_KEY")) 

data.frame(name = c("PAYNSA", "PAYEMS"),
           code = c("CEU0000000001", "CES0000000001")) -> key


dat %>% 
  mutate(date = paste0(periodName, " ", year, " 01"),
         date = as.Date(date, "%B %Y %d")) %>% 
  rename(code = seriesID) %>% 
  left_join(key, by = "code") %>% 
  select(value, date, name) %>% 
  pivot_wider(names_from = "name", values_from = "value") %>% 
  mutate(nfp_chg_sa = PAYEMS - lead(PAYEMS, 1),
         nfp_chg_nsa = PAYNSA - lead(PAYNSA, 1)) -> dat_chg

dat_chg[1:4,] %>% ## edit length of avg here
  select(nfp_chg_nsa, nfp_chg_sa) %>% 
  summarize(avg_sa = mean(nfp_chg_sa),
            avg_nsa = mean(nfp_chg_nsa)) -> avgs_curr_yr

dat_chg[which(dat_chg$date == "2020-02-01"):(which(dat_chg$date == "2020-02-01")+120),] %>% 
  select(date, nfp_chg_nsa, nfp_chg_sa) %>% 
  mutate(month = month(date)) %>% 
  group_by(month) %>% 
  summarize(nfp_chg_sa = mean(nfp_chg_sa),
            nfp_chg_nsa = mean(nfp_chg_nsa)) %>% 
  pivot_longer(-month, names_to = "key", values_to = "avg_10_yr") -> avgs_10_yr


dat_chg %>% 
  select(date, nfp_chg_sa, nfp_chg_nsa) %>% 
  pivot_longer(-date, names_to = "key", values_to = "values") %>% 
  mutate(month = month(date)) %>% 
  left_join(avgs_10_yr, by = c("month", "key")) %>% 
  mutate(avg_curr_yr = case_when(key == "nfp_chg_sa" & date > "2021-08-31" ~ avgs_curr_yr$avg_sa, # edit date
                                 key == "nfp_chg_nsa" & date > "2021-08-31" ~ avgs_curr_yr$avg_nsa)) %>% # edit date
  filter(date > "2021-04-30") -> dat_chg_avg ## edit date

dat_chg_avg %>% 
  select(date, key, values) %>% 
  pivot_wider(everything(), names_from = key, values_from = values) -> dat_chg_wide

dat_chg_avg %>% 
  select(date, key, avg_10_yr) %>% 
  pivot_wider(everything(), names_from = key, values_from = avg_10_yr) %>% 
  rename(avg_10_sa = nfp_chg_sa,
         avg_10_nsa = nfp_chg_nsa) -> avg_10_wide

dat_chg_avg %>% 
  select(date, key, avg_curr_yr) %>% 
  pivot_wider(everything(), names_from = key, values_from = avg_curr_yr) %>% 
  rename(avg_curr_sa = nfp_chg_sa,
         avg_curr_nsa = nfp_chg_nsa) -> avg_curr_wide
  
dat_chg_wide %>% 
  left_join(avg_10_wide, by = c("date")) %>% 
  left_join(avg_curr_wide, by = c("date")) %>% 
  pivot_longer(-date, values_to = "values", names_to = "key") %>% 
  mutate(placeholder = "x") -> dat_final
  
  

ctitle_1 <- "Seasonally and non-seasonally adjusted payroll gains"
unit_label <- "Change in jobs, thousands"
source_1 <- "Source: BLS, @benbakkum."

fmt_dcmls <- function(decimals = 0) {
  function(x) as.character(format(round(as.numeric(x), 0), nsmall=0, big.mark=","))
}


dat_final %>% 
  filter(key %in% c("nfp_chg_sa", "nfp_chg_nsa")) %>% 
  ggplot(aes(date, values, fill = key)) +
  geom_col(position = "dodge", stat = "identity", alpha = 0.8) +
  geom_point(data = dat_final %>% filter(key %in% c("avg_10_sa", "avg_10_nsa")), 
             aes(date, values, shape = key, color = key), 
             position = position_dodge(width = 25), 
             size = 3,
             alpha = 0.6) + 
  geom_point(data = dat_final %>% filter(key %in% c("avg_curr_sa", "avg_curr_nsa")), 
             aes(date, values, shape = key, color = key), 
             position = position_dodge(width = 25), 
             size = 4,
             alpha = 0.6) +
  geom_hline(yintercept = 0, 
             linetype = "dashed", 
             color = cpal("border grey"), 
             size = 1) +
  scale_x_date(date_breaks = "1 month",
               date_labels = "%b '%y") +
  scale_y_continuous(labels = fmt_dcmls(0)) +
  scale_fill_manual(values = c(cpal("blue"), cpal("green"), 
                               cpal("blue"), cpal("green"), 
                               cpal("blue"), cpal("green")),
                    labels = c("Non-seasonally adjusted",
                               "Seasonally adjusted"),
                    breaks = c("nfp_chg_sa", "nfp_chg_nsa")) +
  scale_shape_manual(values = c("avg_10_nsa" = 15, 
                                "avg_10_sa"  = 15,
                                "avg_curr_sa" = 18,
                                "avg_curr_nsa" = 18),
                     labels = c("10y avg for given month (pre-covid)",
                                "Last four month avg"),
                     breaks = c("avg_10_nsa",
                                "avg_curr_sa")) +
  scale_color_manual(values = c("avg_10_nsa" = cpal("purple"), 
                                "avg_10_sa"  = cpal("purple"),
                                "avg_curr_sa" = cpal("orange"),
                                "avg_curr_nsa" = cpal("orange")),
                     labels = c("10y avg for given month (pre-covid)",
                                "Last four month avg"),
                     breaks = c("avg_10_nsa",
                                "avg_curr_nsa")) +
  guides(fill = guide_legend(override.aes = list(fill = c(cpal("blue"), cpal("green")),
                                                 shape = c(NA, NA),
                                                 color = c(NA, NA))),
         ncol = 2,
         bycol = T) +
  my_theme +
  theme(legend.direction = "vertical",
        legend.text = element_text(margin = margin(t = 5)),
        axis.text.x = element_text(size = 14, color = cpal("text"), family = "Open Sans",
                               margin = unit(c(0.35, 0.25, 0.25, 0.25), "cm"),
                               angle = 50, vjust = 0.85, hjust = 0.75)) +
  labs(title = ctitle_1, subtitle = unit_label, caption = source_1) +
  transition_manual(placeholder) -> p

anim1 <- animate(p, renderer = gifski_renderer(loop = T), nframe = 1,
                 width = 615, height = 475, type = "cairo")

anim1
  
```


Despite the lower than expected net number of job gains, the unemployment rate and employment-to-population ratios improved in a way that made it look like a month of strong labor market conditions.

```{r echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE}


dat <- bls_api(c("LNS14000000", "LNS14000060", "LNS11300000", "LNS11300060", "LNS12300000", "LNS12300060", "LNS13327709",
                "LNS12032194", "LNS12000000", "LNS13008636", "LNS11000000", "CES0500000003", "CES0000000001"), 
              startyear = 2018, 
              endyear = 2021, 
              Sys.getenv("BLS_KEY"))

data.frame(name = c("UR", "Prime_age_UR", "LFPR", "Prime_age_LFPR", "EPOP", "Prime_age_EPOP", "U6", 
                    "PTER", "Employment", "Long_term_unemployed", "LF", "AHE", "PAYEMS"),
           code = c("LNS14000000", "LNS14000060", "LNS11300000", "LNS11300060", "LNS12300000", "LNS12300060", "LNS13327709",
                "LNS12032194", "LNS12000000", "LNS13008636", "LNS11000000", "CES0500000003", "CES0000000001")) -> key
```

```{r table, echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE, results='asis'}

dat <- bls_api(c("LNS14000000", "LNS14000060", "LNS11300000", "LNS11300060", "LNS12300000", "LNS12300060", "LNS13327709",
                "LNS12032194", "LNS12000000", "LNS13008636", "LNS11000000", "CES0500000003", "CES0000000001"), 
              startyear = 2018, 
              endyear = 2021, 
              Sys.getenv("BLS_KEY"))

data.frame(name = c("UR", "Prime_age_UR", "LFPR", "Prime_age_LFPR", "EPOP", "Prime_age_EPOP", "U6", 
                    "PTER", "Employment", "Long_term_unemployed", "LF", "AHE", "PAYEMS"),
           code = c("LNS14000000", "LNS14000060", "LNS11300000", "LNS11300060", "LNS12300000", "LNS12300060", "LNS13327709",
                "LNS12032194", "LNS12000000", "LNS13008636", "LNS11000000", "CES0500000003", "CES0000000001")) -> key

dat %>% 
  mutate(date = paste0(periodName, " ", year, " 01"),
         date = as.Date(date, "%B %Y %d")) %>% 
  rename(code = seriesID) %>% 
  left_join(key, by = "code") %>% 
  select(value, date, name) %>% 
  pivot_wider(names_from = "name", values_from = "value") %>% 
  mutate(PTER_emp = round((PTER / Employment) * 100, 1),
         Long_term_UR = round((Long_term_unemployed / LF) * 100, 1),
         AHE_mm = round((AHE / lead(AHE, 1) - 1) * 100, 1),
         AHE_yy = round((AHE / lead(AHE, 12) - 1) * 100, 1),
         nfp_chg = PAYEMS - lead(PAYEMS, 1),
         ) %>% 
  select(date, nfp_chg, UR, Prime_age_UR, U6, LFPR, Prime_age_LFPR, EPOP, Prime_age_EPOP, 
         PTER_emp, Long_term_UR, AHE_mm, AHE_yy) %>% 
  head(11) %>% 
  arrange(date) %>% 
  mutate(date = format(date, "%b-%y")) %>% 
  t() %>% 
  as_tibble(rownames = "rowname") %>% 
  row_to_names(row_number = 1) -> df_tbl

df_tbl[,-1] %>% 
  mutate_if(is.character, as.numeric) %>% 
  bind_cols(df_tbl$date) %>% 
  rename(date = ...12) %>% 
  select(date, everything()) %>% 
  mutate(m1_chg = .[[3]] - .[[2]],
         m2_chg = .[[4]] - .[[3]],
         m3_chg = .[[5]] - .[[4]],
         m4_chg = .[[6]] - .[[5]],
         m5_chg = .[[7]] - .[[6]],
         m6_chg = .[[8]] - .[[7]],
         m7_chg = .[[9]] - .[[8]],
         m8_chg = .[[10]] - .[[9]],
         m9_chg = .[[11]] - .[[10]],
         m10_chg = .[[12]] - .[[11]]) -> df_chgs

df_chgs[,c(1,3,13,4,14,5,15,6,16,7,17,8,18,9,19,10,20,11,21,12,22)] %>% 
  rename(" " = date) %>% 
  as.data.frame() -> df_tbl

## row names
df_tbl[1,1] <- "NFP growth, k"
df_tbl[2,1] <- "UR, %"
df_tbl[3,1] <- "Prime-age UR, %"
df_tbl[4,1] <- "U6, %"
df_tbl[5,1] <- "LFPR, %"
df_tbl[6,1] <- "Prime-age LFPR, %"
df_tbl[7,1] <- "EPOP, %"
df_tbl[8,1] <- "Prime-age EPOP, %"
df_tbl[9,1] <- "PTER, % of payrolls"
df_tbl[10,1] <- "Long-term UR, %"
df_tbl[11,1] <- "AHE, % m/m"
df_tbl[12,1] <- "AHE, % y/y"

df_spk <- df_tbl[,c(1, seq(2,ncol(df_tbl), 2))]


## top row
fmt_dcmls <- function(x) {
   format(round(as.numeric(x), 0), nsmall=0, big.mark=",")
}

for (i in seq(2,(ncol(df_tbl)-1),2)) {
  df_tbl[1,i] <- fmt_dcmls(df_tbl[1,i])
}

hlight <- function(x) {
  y <- as.numeric(gsub(",", "", x))

  case_when(y > 500              ~ color_tile("#75e86d", "#75e86d")(x),
            y < 500 & y > 0      ~ color_tile("#cfffcc", "#cfffcc")(x),
            y < 0   & y > -500   ~ color_tile("#ffd6d6", "#ffd6d6")(x),
            y < -500             ~ color_tile("#f59393", "#f59393")(x))
}

for (i in seq(2,(ncol(df_tbl)-1),2)) {
  df_tbl[1,i] <- hlight(df_tbl[1,i])
}


## all rows
fmt_1dcml <- function(x) {
  y <- format(round(as.numeric(x), 1), nsmall=1)
  x <- as.character(y)
  z <- as.numeric(str_sub(x, 1, 4))
  z
}

for (j in 3:nrow(df_tbl)) {
  for (i in 2:ncol(df_tbl)) {
    df_tbl[j,i] <- fmt_1dcml(df_tbl[j,i])
  }
}

## decline is good
dec_is_good <- function(i, j) {
  y <- as.numeric(df_tbl[j,i+1])

  case_when(y <  -0.1              ~ color_tile("#75e86d", "#75e86d")(df_tbl[j,i]),
            (y >= -0.1) & (y < 0)      ~ color_tile("#cfffcc", "#cfffcc")(df_tbl[j,i]),
            y ==    0              ~ df_tbl[j,i],
            (y <=  0.1) & (y > 0)      ~ color_tile("#ffd6d6", "#ffd6d6")(df_tbl[j,i]),
            y >   0.1              ~ color_tile("#f59393", "#f59393")(df_tbl[j,i]))
}

for (j in c(2, 3, 4, 9, 10)) {
  for (i in seq(2,(ncol(df_tbl)-1),2)) {
    df_tbl[j,i] <- dec_is_good(i, j)
  }
}


## increase is good
inc_is_good <- function(i, j) {
  y <- as.numeric(df_tbl[j,i+1])

  # case_when(y >    0.1              ~ color_tile("#75e86d", "#75e86d")(df_tbl[j,i]),
  #           (y <=   0.1) & (y > 0)      ~ color_tile("#cfffcc", "#cfffcc")(df_tbl[j,i]),
  #           y ==     0              ~ df_tbl[j,i],
  #           (y >=  -0.1) & (y < 0)      ~ color_tile("#ffd6d6", "#ffd6d6")(df_tbl[j,i]),
  #           y <   -0.1              ~ color_tile("#f59393", "#f59393")(df_tbl[j,i]))
  
    ifelse(y >    0.1,              color_tile("#75e86d", "#75e86d")(df_tbl[j,i]),
    ifelse((y <=   0.1) & (y > 0),  color_tile("#cfffcc", "#cfffcc")(df_tbl[j,i]),
    ifelse(y ==     0,             df_tbl[j,i],
    ifelse((y >=  -0.1) & (y < 0),  color_tile("#ffd6d6", "#ffd6d6")(df_tbl[j,i]),
            color_tile("#f59393", "#f59393")(df_tbl[j,i])))))
}

for (j in c(5, 6, 7, 8, 11, 12)) {
  for (i in seq(2,(ncol(df_tbl)-1),2)) {
    df_tbl[j,i] <- inc_is_good(i, j)
  }
}


## https://github.com/haozhu233/kableExtra/issues/217

df_tbl %>% 
  kable(escape = F,
        align = c('l', rep('r', 19))) %>% 
  kable_paper("hover", full_width = F, html_font = "Open Sans") %>% 
  remove_column(3) %>% 
  remove_column(4) %>% 
  remove_column(5) %>% 
  remove_column(6) %>% 
  remove_column(7) %>% 
  remove_column(8) %>% 
  remove_column(9) %>% 
  remove_column(10) %>% 
  remove_column(11) %>% 
  remove_column(12) %>% 
  remove_column(2) %>% 
  remove_column(2) %>% 
  remove_column(2) %>% 
  add_footnote("Source: BLS, @benbakkum.", notation = "none") -> table

table

```
&nbsp;



```{r echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE}
dat %>% 
  mutate(date = paste0(periodName, " ", year, " 01"),
         date = as.Date(date, "%B %Y %d")) %>% 
  rename(code = seriesID) %>% 
  left_join(key, by = "code") %>% 
  select(value, date, name) %>% 
  pivot_wider(names_from = "name", values_from = "value") %>% 
  mutate(PTER_emp = round((PTER / Employment) * 100, 1),
         Long_term_UR = round((Long_term_unemployed / LF) * 100, 1),
         AHE_mm = round((AHE / lead(AHE, 1) - 1) * 100, 1),
         AHE_yy = round((AHE / lead(AHE, 12) - 1) * 100, 1),
         nfp_chg = PAYEMS - lead(PAYEMS, 1),
         ) %>% 
  select(date, nfp_chg, UR, Prime_age_UR, U6, LFPR, Prime_age_LFPR, EPOP, Prime_age_EPOP, 
         PTER_emp, Long_term_UR, AHE_mm, AHE_yy) -> chrt_dat


chrt_dat %>% 
  pivot_longer(!date, names_to = "key", values_to = "values") %>% 
  filter(date >= as.Date("2018-12-31")) %>% 
  mutate(key = case_when(key == "nfp_chg"         ~ "NFP growth, k",
                         key == "UR"              ~ "UR, %",
                         key == "Prime_age_UR"    ~ "Prime-age UR, %",
                         key == "U6"              ~ "U6, %",
                         key == "LFPR"            ~ "LFPR, %",
                         key == "Prime_age_LFPR"  ~ "Prime-age LFPR, %",
                         key == "EPOP"            ~ "EPOP, %",
                         key == "Prime_age_EPOP"  ~ "Prime-age EPOP, %",
                         key == "PTER_emp"        ~ "PTER, % of payrolls",
                         key == "Long_term_UR"    ~ "Long-term UR, %",
                         key == "AHE_mm"          ~ "AHE, % m/m",
                         key == "AHE_yy"          ~ "AHE, % y/y")) -> chrt_dat_lng

chrt_dat_lng$key <- factor(chrt_dat_lng$key, 
                           levels = c("NFP growth, k",
                         "UR, %",
                         "Prime-age UR, %",
                         "U6, %",
                         "LFPR, %",
                         "Prime-age LFPR, %",
                         "EPOP, %",
                         "Prime-age EPOP, %",
                         "PTER, % of payrolls",
                         "Long-term UR, %",
                         "AHE, % m/m",
                         "AHE, % y/y"))      


source_1 <- "Source: BLS, @benbakkum."

chart1 <- chrt_dat_lng %>% 
  ggplot(aes(date, values)) +
  geom_step(color = cpal("blue")) +
  scale_x_date(breaks = c(as.Date("2019-01-01"), as.Date("2020-01-01"), as.Date("2021-01-01")),
               date_labels = "'%y") +
  facet_wrap(~key, scales = "free_y", ncol = 4) +
  theme(
    legend.position = "none",
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = cpal("light grey"), size = 0.21),
    panel.border = element_rect(fill = NA, color = cpal("border grey"), size = 0.3),
    plot.title = element_text(size = 21, color = "black", family = "Open Sans Light"),
    plot.subtitle = element_text(size = 14, color = cpal("text"), family = "Open Sans"),
    plot.caption = element_text(size = 12.5, color = cpal("text"), family = "Open Sans"),
    text = element_text(size = 15, color = cpal("text"), family = "Open Sans"),
    legend.text = element_text(size = 15, color = cpal("text"), family = "Open Sans"),
    axis.text.x = element_text(size = 14, color = cpal("text"), family = "Open Sans",
                               margin = unit(c(0.35, 0.25, 0.25, 0.25), "cm")),
    axis.text.y = element_text(size = 14, color = cpal("text"), family = "Open Sans",
                               margin = unit(c(0.35, 0.25, 0.25, 0.25), "cm")),
    axis.title.y = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.length = unit(-0.15, "cm"),
    axis.ticks = element_line(color = cpal("border grey"), size = 0.3),
    plot.margin = unit(c(0.3, 0.5, 0.3, 0.05), "cm"),
    strip.text.x = element_text(size = 12, color = cpal("text"), family = "Open Sans")
  ) +
  labs(caption = source_1) +
  transition_reveal(date)
  
  
anim1 <- animate(chart1, renderer = gifski_renderer(loop = T), nframe = 225,
                 width = 615, height = 500, type = "cairo", end_pause = 200)  

anim1
```

Transportation and warehousing as well as professional service jobs continue to stand out as industries where employment is growing at a strong pace, even though they're already well above pre-pandemic levels of employment. Leisure and hospitality as well as health care job gains have slowed, likely reflecting the ongoing effects of the pandemic. 

```{r industries, echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE}
codes <- c("USCONS", "MANEMP", "USMINE", "USWTRADE", "USTRADE", "CES4300000001", "CES4422000001", "CES6054000001",
           "USINFO", "USFIRE", "CES6056000001", "CES6561000001", "CES6562000001", "USLAH", "USSERV")
key <- c("Construction", "Manufacturing", "Mining & Logging", "Wholesale Trade", "Retail Trade",
           "Transport & Warehouse", "Utilities", "Professional Svcs", "Information", "Financial",
           "Admin & Waste Svcs", "Educational Svcs", "Health Care",
           "Leisure & Hospitality", "Other Services")
cat <- c("Goods-producing", "Goods-producing", "Goods-producing", "Service-providing", "Service-providing",
         "Service-providing", "Service-providing", "Service-providing", "Service-providing", "Service-providing",
         "Service-providing", "Service-providing", "Service-providing", "Service-providing", "Service-providing")
cat_map <- data.frame(key, cat)

raw_dat <- getSymbols(codes,src='FRED') 

dat <- merge(USCONS, MANEMP, USMINE, USWTRADE, USTRADE, CES4300000001, CES4422000001, CES6054000001,
           USINFO, USFIRE, CES6056000001, CES6561000001, CES6562000001, USLAH, USSERV)  %>% 
  as_tibble(rownames = NA) %>% 
  rownames_to_column() 

colnames(dat) <- c("date", key)


chart1 <- dat %>% 
  filter(date > "2018-12-31") %>% 
  pivot_longer(!date) %>% 
  filter(!name %in% c("Other Services", "Utilities", "Wholesale Trade")) %>% 
  mutate(date = as.Date(date, "%Y-%m-%d"),
         name = fct_reorder(name, value, function(.) min(.) / max(.))) %>% 
  ggplot(aes(date, value)) +
  geom_step(color = cpal("blue")) +
  scale_x_date(breaks = c(as.Date("2019-01-01"), as.Date("2020-01-01"), as.Date("2021-01-01")),
               date_labels = "'%y") +
  facet_wrap(~ name, scales = "free_y") +
   theme(
    legend.position = "none",
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = cpal("light grey"), size = 0.21),
    panel.border = element_rect(fill = NA, color = cpal("border grey"), size = 0.3),
    plot.title = element_text(size = 21, color = "black", family = "Open Sans Light"),
    plot.subtitle = element_text(size = 14, color = cpal("text"), family = "Open Sans"),
    plot.caption = element_text(size = 12.5, color = cpal("text"), family = "Open Sans"),
    text = element_text(size = 15, color = cpal("text"), family = "Open Sans"),
    legend.text = element_text(size = 15, color = cpal("text"), family = "Open Sans"),
    axis.text.x = element_text(size = 14, color = cpal("text"), family = "Open Sans",
                               margin = unit(c(0.35, 0.25, 0.25, 0.25), "cm")),
    axis.text.y = element_text(size = 14, color = cpal("text"), family = "Open Sans",
                               margin = unit(c(0.35, 0.25, 0.25, 0.25), "cm")),
    axis.title.y = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.length = unit(-0.15, "cm"),
    axis.ticks = element_line(color = cpal("border grey"), size = 0.3),
    plot.margin = unit(c(0.3, 0.5, 0.3, 0.05), "cm"),
    strip.text.x = element_text(size = 11, color = cpal("text"), family = "Open Sans")
  ) +
  labs(caption = source_1) +
  transition_reveal(date)

anim1 <- animate(chart1, renderer = gifski_renderer(loop = T), nframe = 225,
                 width = 615, height = 500, type = "cairo", end_pause = 200)  

anim1

```

The chart below shows the strong pace of this recovery relative to other recessions.

```{r PAYEMS, echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE}


raw_dat <- getSymbols('PAYEMS',src='FRED', auto.assign=F) 

dat <- raw_dat %>% 
  as_tibble(rownames = NA) %>% 
  rownames_to_column() %>% 
  rename(date = rowname,
         jobs = PAYEMS) %>% 
  mutate(date = as.Date(date, "%Y-%m-%d"))

dat_ph <- dat %>% 
  mutate(recess = case_when(date >= "1948-09-01" & date <= "1950-07-01" ~ "1948", 
                            date >= "1953-07-01" & date <= "1955-06-01" ~ "1953",
                            date >= "1957-04-01" & date <= "1959-04-01" ~ "1957",
                            date >= "1960-04-01" & date <= "1961-12-01" ~ "1960",
                            date >= "1974-10-01" & date <= "1976-02-01" ~ "1974",
                            date >= "1980-03-01" & date <= "1981-01-01" ~ "1980",
                            date >= "1981-07-01" & date <= "1983-11-01" ~ "1981",
                            date >= "1990-06-01" & date <= "1993-02-01" ~ "1990",
                            date >= "2001-02-01" & date <= "2005-02-01" ~ "2001",
                            date >= "2008-01-01" & date <= "2014-05-01" ~ "2008",
                            date >= "2020-02-01" ~ "2020"))

dat_long <- dat_ph %>% 
  group_by(recess) %>% 
  mutate(perc = (jobs / first(jobs) - 1) * 100) %>% 
  na.omit() %>% 
  mutate(counter = row_number() - 1) %>%
  ungroup() %>% 
  mutate(idx = ifelse(recess == "1948", counter, #update date here
               ifelse(recess == "1953", counter + 4,
               ifelse(recess == "1957", counter + 8,
               ifelse(recess == "1960", counter + 12,
               ifelse(recess == "1974", counter + 16,
               ifelse(recess == "1980", counter + 20,
               ifelse(recess == "1981", counter + 24,
               ifelse(recess == "1990", counter + 28,
               ifelse(recess == "2001", counter + 32,
               ifelse(recess == "2008", counter + 36,
               ifelse(recess == "2020" & date == "2020-02-01", counter + 106,
               ifelse(recess == "2020" & date == "2020-03-01", counter + 110,
               ifelse(recess == "2020" & date == "2020-04-01", counter + 130,
               ifelse(recess == "2020" & date == "2020-05-01", counter + 132,
               ifelse(recess == "2020" & date == "2020-06-01", counter + 134,
               ifelse(recess == "2020" & date == "2020-07-01", counter + 138, 
               ifelse(recess == "2020" & date == "2020-08-01", counter + 142, 
               ifelse(recess == "2020" & date == "2020-09-01", counter + 144, 
               ifelse(recess == "2020" & date == "2020-10-01", counter + 146, 
               ifelse(recess == "2020" & date == "2020-11-01", counter + 148, 
               ifelse(recess == "2020" & date == "2020-12-01", counter + 150, 
               ifelse(recess == "2020" & date == "2021-01-01", counter + 152, 
               ifelse(recess == "2020" & date == "2021-02-01", counter + 154,
               ifelse(recess == "2020" & date == "2021-03-01", counter + 156,
               ifelse(recess == "2020" & date == "2021-04-01", counter + 158,
               ifelse(recess == "2020" & date == "2021-05-01", counter + 160,
               ifelse(recess == "2020" & date == "2021-06-01", counter + 162, 
               ifelse(recess == "2020" & date == "2021-07-01", counter + 164,
               ifelse(recess == "2020" & date == "2021-08-01", counter + 166, 
               ifelse(recess == "2020" & date == "2021-09-01", counter + 168, 
               ifelse(recess == "2020" & date == "2021-10-01", counter + 170, 
               ifelse(recess == "2020" & date == "2021-11-01", counter + 172,
               ifelse(recess == "2020" & date == "2021-12-01", counter + 174, NA)))))))))))))))))))))))))))))))))) %>%
  mutate(recess = paste0(recess, " "))
        

start_date <- head(dat_long$date,1)
end_date <- tail(dat_long$date,1)

line_cols <- c("1948 " = "#b2df8a", 
               "1953 " = "#fb9a99", 
               "1957 " = "#fdbf6f", 
               "1960 " = "#ff7f00", 
               "1974 " = "#cab2d6", 
               "1980 " = "#6a3d9a", 
               "1981 " = "#b15928", 
               "1990 " = "#33a02c",
               "2001 " = "#a6cee3", 
               "2008 " = "#1f78b4", 
               "2020 " = "#e31a1c")

fmt_dcmls <- function(decimals = 0) {
  function(x) as.character(format(round(as.numeric(x), 0), nsmall=0, big.mark=","))
}

ctitle_1 <- "Job losses during US post-war recessions"
unit_label <- "% difference from cycle's peak total number of jobs"
source_1 <- "Source: BLS, FRED, @benbakkum."
x_lab <- "Number of months after peak employment"

chart_1 <- dat_long %>%
  ggplot(aes(x = counter, y = perc, color = recess)) +
  geom_line(size = 1, alpha = 0.85) +
  geom_point(size = 2.2) +
  geom_text(data = filter(dat_long, recess == "2008 " | recess == "2020 " | recess == "2001 "), 
            aes(x = counter, y = perc, label = recess),
            nudge_x = 3.2, 
            nudge_y = 0.15,
            show.legend = FALSE, 
            size = 4, 
            family = "Open Sans", 
            fontface = "bold") +
  scale_color_manual(values = line_cols) +
  scale_x_continuous(limits = c(0, 85), expand = c(0,0)) +
  scale_y_continuous(limits = c(-16, 1), expand = c(0,0), labels = fmt_dcmls(0)) +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.border = element_rect(fill = NA, color = cpal("border grey"), size = 0.3),
    plot.title = element_text(size = 21, color = "black", family = "Open Sans Light"),
    plot.subtitle = element_text(size = 14, color = cpal("text"), family = "Open Sans"),
    plot.caption = element_text(size = 12.5, color = cpal("text"), family = "Open Sans"),
    text = element_text(size = 15, color = cpal("text"), family = "Open Sans"),
    legend.text = element_text(size = 15, color = cpal("text"), family = "Open Sans"),
    axis.text.x = element_text(size = 14, color = cpal("text"), family = "Open Sans",
                               margin = unit(c(0.35, 0.25, 0.25, 0.25), "cm")),
    axis.text.y = element_text(size = 14, color = cpal("text"), family = "Open Sans",
                               margin = unit(c(0.25, 0.25, 0.25, 0.25), "cm")),
    axis.title.y = element_blank(),
    axis.title.x = element_text(size = 14, color = cpal("text"), family = "Open Sans"),
    axis.ticks.length = unit(-0.15, "cm"),
    axis.ticks = element_line(color = cpal("border grey"), size = 0.3),
    panel.grid.major = element_line(color = cpal("light grey"), size = 0.21),
    plot.margin = unit(c(0.3, 0.5, 0.3, 0.3), "cm"),
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.key = element_blank(),
    legend.key.size = unit(0.37, "cm"),
    legend.background = element_blank(),
    legend.margin = margin(t = 0, unit = "cm"),
    legend.spacing.x = unit(0.2, 'cm'),
    legend.spacing.y = unit(0.2, 'cm')) +
  guides(color=guide_legend(nrow=2,byrow=TRUE)) +
  labs(title = ctitle_1, 
       subtitle = unit_label,
       caption = paste("Data as of ", format(tail(dat$date,1), "%b %Y"), ". ", source_1[1], sep = ""),
       x = x_lab) +
  transition_reveal(idx, keep_last = TRUE) +
  ease_aes('linear') +
  view_follow(fixed_x = TRUE)

anim <- animate(chart_1, renderer = gifski_renderer(loop = T), nframe = 360,
                 width = 615, height = 505, type = "cairo", end_pause = 120)

anim
```

If the average job growth of the last three months continues at the same pace, the total number of jobs would fully recover to where it would have likely ended up had the pandemic not occurred around the end of 2024. In many ways that would be an even tighter labor market than before the pandemic as the population has aged and a large number of workers have retired in the intervening period.

```{r echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE}

raw_dat <- bls_api(c("CEU0000000001", "CES0000000001"), 
              startyear = 2005, 
              endyear = 2021,
              Sys.getenv("BLS_KEY")) 

data.frame(name = c("PAYNSA", "PAYEMS"),
           code = c("CEU0000000001", "CES0000000001")) -> key


raw_dat %>% 
  mutate(date = paste0(periodName, " ", year, " 01"),
         date = as.Date(date, "%B %Y %d")) %>% 
  rename(code = seriesID) %>% 
  left_join(key, by = "code") %>% 
  select(value, date, name) %>% 
  pivot_wider(names_from = "name", values_from = "value") %>% 
  select(-PAYNSA) %>% 
  arrange(date) %>% 
  mutate(since_feb20 = case_when(date == ymd("2020-02-01") ~ PAYEMS,
                                 date != ymd("2020-02-01") ~ NA_real_),
         since_now =   case_when(date == ymd("2021-12-01") ~ PAYEMS, #edit this
                                 date != ymd("2021-12-01") ~ NA_real_)) -> dat #edit this


data.frame(date = month_dates <- seq(last(dat$date,1) %m+% months(1), last(dat$date,1) %m+% months(48), by = "month"),
           PAYEMS = rep(NA, length(month_dates)),
           since_feb20 = rep(NA, length(month_dates)),
           since_now = rep(NA, length(month_dates))) -> future

dat %>% 
  rbind(future) -> dat_fut

row_start_feb20 <- which(grepl(ymd("2020-02-01"), dat_fut$date))
row_start_now <- which(grepl(ymd("2021-12-01"), dat_fut$date)) #edit this


dat_chg[which(dat_chg$date == "2020-02-01"):(which(dat_chg$date == "2020-02-01")+120),] %>% 
  select(date, nfp_chg_nsa, nfp_chg_sa) %>% 
  summarize(avg = mean(nfp_chg_sa)) %>% 
  unlist() -> pre_covid_10y_avg


dat_chg[1:3,] %>% 
  select(nfp_chg_sa) %>% 
  summarize(avg_sa = mean(nfp_chg_sa)) %>% 
  unlist() -> last_mths_avg


dat_fut %>% 
  mutate(since_feb20 = case_when(date < ymd("2020-02-01") ~ PAYEMS,
                                 date >= ymd("2020-02-01") ~ pull(dat[row_start_feb20,"since_feb20"]) + 
                                                          (pre_covid_10y_avg * (row_number() - row_start_feb20))),
         since_now = case_when(date < ymd("2021-12-01") ~ PAYEMS, #edit this
                              date >= ymd("2021-12-01") ~ pull(dat[row_start_now,"since_now"]) + #edit this
                                                          (last_mths_avg * (row_number() - row_start_now)))) -> df_final



ctitle_1 <- "Nonfarm payrolls w/ current trajectory extrapolated"
unit_label <- "Payrolls, thousands, seasonally adjusted"
source_1 <- "Source: BLS, @benbakkum."

fmt_dcmls <- function(decimals = 0) {
  function(x) as.character(format(round(as.numeric(x), 0), nsmall=0, big.mark=","))
}

df_final %>% 
  pivot_longer(-date, names_to = "key", values_to = "values") %>% 
  mutate(key = case_when(key == "PAYEMS" ~ "Actual total jobs",
                         key == "since_feb20" ~ "Pre-pandemic trend",
                         key == "since_now" ~ "Last three months trend")) -> df_long #edit this

df_long$key <- factor(df_long$key, c("Pre-pandemic trend",
                                     "Last three months trend",
                                     "Actual total jobs"
                                     ))
                                     
                                     

df_long %>% 
  filter(date > "2017-12-01" & date < "2024-12-01") %>% 
  ggplot(aes(date, values, color = key, linetype = key, size = key)) +
  geom_line() +
  geom_point(size = 2.5) +
  scale_linetype_manual(values = c("dashed", "dashed", "solid")) +
  scale_color_manual(values = c(cpal("green"), cpal("orange"), cpal("blue"))) +
  scale_size_manual(values = c(1, 1, 1.2)) +
  scale_x_date(expand = c(0,0),
               limits = c(ymd("2019-01-01"), ymd("2024-12-01")),
               date_labels = "'%y",
               date_breaks = "1 year") +
  scale_y_continuous(labels = fmt_dcmls(0)) +
  labs(title = ctitle_1, subtitle = unit_label, caption = source_1) +
  my_theme +
  guides(color=guide_legend(ncol=1),
         linetype=guide_legend(ncol=1)) +
  theme(legend.key.height = unit(0.7, "cm")) +
  transition_reveal(date, keep_last = TRUE) -> chart1




anim1 <- animate(chart1, renderer = gifski_renderer(loop = T), nframe = 300,
                 width = 615, height = 460, type = "cairo", end_pause = 200)

anim1
```


An aggregate measure of labor market conditions, the Blanchflower-Levin employment gap, shows that slack continues to decline, with underemployment (a major source of slack during the last cycle) and unemployment slack having already disappeared according to this measure. My interest continues to be in how long the participation gap will linger during this cycle.


```{r blanchflower, echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE}
library(ggforce)
library(lubridate)

raw_dat <- getSymbols(c('UNRATE', 'CIVPART', 'LNS12032194', 'CLF16OV') ,src='FRED', auto.assign=T) 

dat <- merge(UNRATE, CIVPART, LNS12032194, CLF16OV) %>% 
  as_tibble(rownames = NA) %>% 
  rownames_to_column() %>% 
  rename(date = rowname,
         UR = UNRATE,
         LFPR = CIVPART,
         PTER = LNS12032194,
         LF = CLF16OV) %>% 
  mutate(date = as.Date(date, "%Y-%m-%d"),
         date = ceiling_date(date, "month") - 1)

## NAIRU
raw_nairu <- getSymbols('NROU' ,src='FRED', auto.assign=T) 

dat_nairu <- NROU %>% 
  as_tibble(rownames = NA) %>% 
  rownames_to_column() %>% 
  rename(date = rowname,
         NAIRU = NROU) %>% 
  mutate(date = as.Date(date, "%Y-%m-%d")) %>% 
  filter(date < Sys.Date())


mthly_nairu <- seq(dat_nairu$date[1] + 1, 
                   tail(dat_nairu$date,1) + 1, 
                   by = "1 month") - 1

nairu_m <- data.frame(date = ceiling_date(mthly_nairu, "month") - 1, 
                      nairu = spline(dat_nairu, 
                                      method = "fmm", 
                                      xout=mthly_nairu)$y) %>% 
  as_data_frame()
  


## potential LFPR
cbo_dat <- read_csv("D:\\Documents\\R directory\\cbo_dat.csv") %>% 
  select(date, potential_lfpr) %>% 
  mutate(date = as.yearqtr(date, "%Yq%q"),
         date = as.Date(date, 1))
  
monthly <- seq(cbo_dat$date[1] + 1, tail(cbo_dat$date,1) + 1, by = "1 month") - 1
cbo_m <- data.frame(date = monthly, 
                    cbo_lfpr = spline(cbo_dat, 
                                      method = "fmm", 
                                      xout=monthly)$y) %>% 
  as_data_frame() %>% 
  filter(date < Sys.Date())


## merge and gaps calculated
dat_tot <- dat %>% 
  left_join(nairu_m, by = "date") %>% 
  fill(nairu) %>% 
  left_join(cbo_m, by = "date") %>% 
  filter(date > "1979-12-31",
         date <= tail(dat$date,1)) %>% 
  mutate(PTER_adj = PTER * (5.48/8.08),
         PTER_perc = PTER_adj / LF)

pter_avg <- dat_tot %>% 
  select(date, PTER_perc) %>% 
  filter(date > "1993-12-31",
         date <= "2007-12-31") %>% 
  summarize(avg = mean(PTER_perc)) %>% 
  unlist()

BL <- dat_tot %>% 
  mutate(PTER_gap = (PTER_perc - pter_avg) * 100,
         UR_gap = UR - nairu,
         LFPR_gap = cbo_lfpr - LFPR,
         total = PTER_gap + UR_gap + LFPR_gap) %>% 
  select(date, PTER_gap, UR_gap, LFPR_gap) %>%
  rename('Underemployment gap' = PTER_gap, 'Unemployment gap' = UR_gap, 'Pariticipation gap' = LFPR_gap) %>% 
  pivot_longer(cols = ends_with("gap"), names_to = "key", values_to = "value")

dat_line <- dat_tot %>% 
  mutate(PTER_gap = (PTER_perc - pter_avg) * 100,
         UR_gap = UR - nairu,
         LFPR_gap = cbo_lfpr - LFPR,
         total = PTER_gap + UR_gap + LFPR_gap)

ctitle_1 <- "Blanchflower-Levin employment gap"
unit_label <- "%"
source_1 <- "Source: BLS, FRED, @benbakkum."

start_date <- head(dat_line$date,1)
end_date <- tail(dat$date,1)

BL_chart <- BL %>% 
  mutate(placeholder = "X") %>% 
  ggplot() +
  geom_bar(aes(fill = key, y = value, x = date),
           position = "stack", 
           stat = "identity",
           width = 32) +
  facet_zoom(xy = date > "2018/8/1" & date < "2022/01/01", #edit this
             zoom.size = .4) +
  geom_line(data = dat_line, aes(x = date, y = total), 
             size = 0.8, 
             alpha = .5, 
             linetype = "dashed") +
  scale_fill_manual(values = c("#33a02c", "#a6cee3", "#1f78b4")) +
  scale_y_continuous(expand = c(0,0)) +
  scale_x_date(date_labels = "%Y",
               expand = c(0,50)) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.key = element_blank(),
    legend.key.size = unit(0.37, "cm"),
    legend.background = element_blank(),
    legend.margin = margin(t = -0.5, unit = "cm"),
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = cpal("light grey"), size = 0.21),
    panel.border = element_rect(fill = NA, color = cpal("border grey"), size = 0.3),
    plot.title = element_text(size = 21, color = "black", family = "Open Sans Light"),
    plot.subtitle = element_text(size = 14, color = cpal("text"), family = "Open Sans"),
    plot.caption = element_text(size = 12.5, color = cpal("text"), family = "Open Sans"),
    text = element_text(size = 15, color = cpal("text"), family = "Open Sans"),
    legend.text = element_text(size = 12, color = cpal("text"), family = "Open Sans"),
    axis.text.x = element_text(size = 14, color = cpal("text"), family = "Open Sans",
                               margin = unit(c(0.35, 0.25, 0.25, 0.25), "cm")),
    axis.text.y = element_text(size = 14, color = cpal("text"), family = "Open Sans",
                               margin = unit(c(0.35, 0.25, 0.25, 0.25), "cm")),
    axis.title.y = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.length = unit(-0.15, "cm"),
    axis.ticks = element_line(color = cpal("border grey"), size = 0.3),
    plot.margin = unit(c(0.3, 0.5, 0.3, 0.05), "cm")
  ) + 
  labs(title = ctitle_1, subtitle = unit_label,
       caption = paste("Data as of ", format(tail(dat_line$date,1), "%b, %Y"), ". ", 
                       source_1[1], sep = "")) +
  transition_states(placeholder)

anim <- animate(BL_chart, renderer = gifski_renderer(loop = T), fps = 1, duration = 1,
                 width = 615, height = 420, type = "cairo")

anim
```

Note: the unemployment gap is the difference between the unemployment rate and the non-accelerating inflation rate of unemployment (NAIRU). The participation gap is the difference between the labor force participation rate (LFPR) and the CBO’s estimate of the potential LFPR. The underemployment gap is the difference between the number of employees working part-time for economic reasons as a percentage of the labor force, adjusted for the difference in the average number of hours worked by part-time and full-time employees, and the 1994-2007 average of this calculation.





