---
title: "Prices -> Services inflation continuing to accelerate"
date: 2022-01-12T21:13:14-05:00
draft: false
editor_options:
  chunk_output_type: console
---
&nbsp;
&nbsp;

The US Consumer Price Index numbers for December were fairly in line with expectations while at multi-decade highs, but core CPI was a bit hotter than expected. Headline and core CPI came in at 7.0% and 5.5% respectively.

```{r setup, include=FALSE}
library(tidyverse)
library(blscrapeR)
library(tibbletime)
library(lubridate)
library(quantmod)
library(gganimate)
library(ggrepel)
library(purrr)
library(Cairo)
library(magick)
library(gifski)
library(scales)
library(matrixStats)
library(kableExtra)
library(formattable)

windowsFonts("Open Sans" = windowsFont("Open Sans"))
windowsFonts("Open Sans Light" = windowsFont("Open Sans Light"))

colors <- c('paleblue' = "#a6cee3", 'blue' = "#1f78b4", 'green' = "#33a02c", 'palegreen' = "#b2df8a",
            'red' = "#e31a1c", 'palered' = "#fb9a99", 'orange' = "#ff7f00", 'paleorange' = "#fdbf6f",
            'purple' = "#6a3d9a", 'palepurple' = "#cab2d6", 'border grey' = "#808080", 'brown' = '#b15928',
            'text' = "#222222" , 'light grey' = "#CDCDCD")

cpal <- function(...) {
  cols <- c(...)
  if (is.null(cols))
    return(colors)
  colors[[cols]]
}


my_theme <- theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.key = element_blank(),
    legend.key.size = unit(0.37, "cm"),
    legend.background = element_blank(),
    legend.margin = margin(t = -0.5, unit = "cm"),
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = cpal("light grey"), size = 0.21),
    panel.border = element_rect(fill = NA, color = cpal("border grey"), size = 0.3),
    plot.title = element_text(size = 21, color = "black", family = "Open Sans Light"),
    plot.subtitle = element_text(size = 14, color = cpal("text"), family = "Open Sans"),
    plot.caption = element_text(size = 12.5, color = cpal("text"), family = "Open Sans"),
    text = element_text(size = 15, color = cpal("text"), family = "Open Sans"),
    legend.text = element_text(size = 14, color = cpal("text"), family = "Open Sans"),
    axis.text.x = element_text(size = 14, color = cpal("text"), family = "Open Sans",
                               margin = unit(c(0.35, 0.25, 0.25, 0.25), "cm")),
    axis.text.y = element_text(size = 14, color = cpal("text"), family = "Open Sans",
                               margin = unit(c(0.35, 0.25, 0.25, 0.25), "cm")),
    axis.title.y = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.length = unit(-0.15, "cm"),
    axis.ticks = element_line(color = cpal("border grey"), size = 0.3),
    plot.margin = unit(c(0.3, 0.5, 0.3, 0.05), "cm")
  )

knitr::opts_chunk$set(echo = TRUE)

```



```{r echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE}
raw_dat <- getSymbols(c("TRMMEANCPIM159SFRBCLE", "CPILFESL", "MEDCPIM159SFRBCLE"),src='FRED', auto.assign=T) 

key <- c("Trimmed mean", "Core", "Median")


merge(TRMMEANCPIM159SFRBCLE, CPILFESL, MEDCPIM159SFRBCLE)  %>% 
  as_tibble(rownames = NA) %>% 
  rownames_to_column() %>% 
  set_names("date", key) %>% 
  mutate(date = as.Date(date, "%Y-%m-%d"),
         Core = (Core / lag(Core, 12) - 1) * 100) %>% 
  filter(date > "2011-12-31") %>% 
  pivot_longer(-date, names_to = "key", values_to = "values") -> dat



ctitle_1 <- "Measures of CPI central tendency"
unit_label <- "%, year-over-year"
source_1 <- "Source: Cleveland Fed, BLS, FRED, @benbakkum."


dat %>% 
  mutate(placeholder = "x") %>% 
  ggplot(aes(date, values, color = key, linetype = key)) +
  geom_line(size = 1.2, show.legend = F) +
  scale_x_date(expand = c(0,0), 
                     limits = c(as.Date("2012-01-01"), as.Date(tail(dat,1)$date %m+% months(12)))) +
  scale_color_manual(values = c(cpal("blue"), cpal("green"), cpal("orange"))) +
  geom_point(data = tail(dat,3),
             aes(date, values),
             size = 3) +
  geom_text(data = tail(dat,3),
            aes(date, values, label = format(round(values,1), digits = 2)),
            hjust = -.5,
            size = 6,
            fontface = "bold",
            show.legend = F) +
  my_theme +
  labs(title = ctitle_1, subtitle = unit_label, caption = source_1) +
  guides(colour = guide_legend(override.aes = list(size=7))) +
  transition_manual(placeholder) -> p

anim1 <- animate(p, renderer = gifski_renderer(loop = T), nframe = 1,
                 width = 615, height = 420, type = "cairo")
```

Year-over-year rates of inflation can be contextualized with the path of the index and price levels that underlie them. A dip in price levels and a later rebound creates a "base effect" boosting the year-over-year measure of change, even when the index resumes something close to its original trajectory. The beginning of the pandemic certainly saw a drop in prices, but the surge in 2021 far surpassed even where index levels would have been expected to end up had they hypothetically maintained a steady 2% rate of increase over the last year and a half.

```{r echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE}
raw_dat <- getSymbols(c("CPIAUCSL", "CPILFESL"),src='FRED', auto.assign=T) 

key <- c("Headline", "Core")


merge(CPIAUCSL, CPILFESL)  %>% 
  as_tibble(rownames = NA) %>% 
  rownames_to_column() %>% 
  set_names("date", key) %>% 
  mutate(date = as.Date(date, "%Y-%m-%d")) %>% 
  filter(date > "2015-12-31") %>% 
  mutate(Headline = Headline / head(Headline,1) * 100,
         Core = Core / head(Core,1) * 100) %>% 
  mutate(headline_since_19 = case_when(date == ymd("2020-01-01") ~ Headline,
                              date != ymd("2020-01-01") ~ NA_real_),
         core_since_19 = case_when(date == ymd("2020-01-01") ~ Core,
                              date != ymd("2020-01-01") ~ NA_real_)) -> dat
    

mult_factor <- (1.02)^(1/12)

row_start_20 <- which(grepl(ymd("2020-01-01"), dat$date))

dat %>% 
  mutate(headline_since_19 = case_when(date < ymd("2020-01-01") ~ Headline,
                              date >= ymd("2020-01-01") ~ pull(dat[row_start_20,"headline_since_19"]) * 
                                                          (mult_factor ^ (row_number() - row_start_20))),
         core_since_19 = case_when(date < ymd("2020-01-01") ~ Core,
                              date >= ymd("2020-01-01") ~ pull(dat[row_start_20,"core_since_19"]) * 
                                                          (mult_factor ^ (row_number() - row_start_20)))) -> dat_final

ctitle_1 <- "Headline and Core CPI w/ 2% target increases"
unit_label <- "1/1/2016 = 100, seasonally adjusted"
source_1 <- "Source: BLS, FRED, @benbakkum."

fmt_dcmls <- function(decimals = 0) {
  function(x) as.character(format(round(as.numeric(x), 0), nsmall=0, big.mark=","))
}


dat_final %>% 
  pivot_longer(-date, names_to = "key", values_to = "values") %>% 
  mutate(type = case_when(key == "Core" | key == "core_since_19" ~ "Core",
                          key == "Headline" | key == "headline_since_19" ~ "Headline"),
         key = case_when(key == "Core" ~ "Actual Core",
                         key == "Headline" ~ "Actual Headline",
                         key == "headline_since_19" ~ "Headline w/ 2% inflation since '19",
                         key == "core_since_19" ~ "Core w/ 2% inflation since '19")) -> df_long

df_long$key <- factor(df_long$key, c("Headline w/ 2% inflation since '19",
                                     "Actual Headline",
                                     "Core w/ 2% inflation since '19",
                                     "Actual Core"))

df_long$type <- factor(df_long$type, c("Headline", "Core"))

df_long %>% 
  ggplot(aes(date, values, color = key, linetype = key, size = key)) +
  geom_line(show.legend = F) +
  geom_point(size = 2.5) +
  scale_linetype_manual(values = c("dashed", "solid", "dashed", "solid")) +
  scale_color_manual(values = c(cpal("palegreen"), cpal("green"), cpal("paleblue"), cpal("blue"))) +
  scale_size_manual(values = c(0.9, 1.2, 0.9, 1.2)) +
  scale_x_date(expand = c(0,0),
               limits = c(ymd("2016-01-01"), ymd("2022-04-30")),
               date_labels = "'%y",
               date_breaks = "1 year") +
  scale_y_continuous(limits = c(98, 118), expand = c(.02,.02), labels = fmt_dcmls(0),
                     breaks = seq(90, 116, 5)) +
  labs(title = ctitle_1, subtitle = unit_label, caption = source_1) +
  my_theme +
  guides(color=guide_legend(ncol=1, override.aes = list(size=9)), 
         linetype=guide_legend(ncol=2)) +
  theme(legend.key.height = unit(0.7, "cm"),
        strip.text = element_text(size = 14)) +
  facet_wrap(~type) +
  transition_reveal(date, keep_last = TRUE) +
  view_follow(fixed_x = TRUE, fixed_y = FALSE) -> chart1




anim1 <- animate(chart1, renderer = gifski_renderer(loop = T), nframe = 235,
                 width = 615, height = 450, type = "cairo", end_pause = 200)

anim1

```

I also like looking at the distributions of components within CPI because it's fun to watch, but also because it gives an idea of the level of dispersion between inflation for things as varied as men's sweaters and medical services. In the chart below, I take 180 non-overlapping components of CPI and show the distribution of their year-over-year changes over time. Dispersion clearly increased with the onset of the pandemic and then amidst reopening, and since the summer the distribution has shifted to the right. It's striking how few components' prices are declining on a year-over-year basis right now.

```{r echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE}
raw_dat <- read_csv("D://Documents//R directory//inflation//cpi_components.csv")

raw_dat %>% 
  filter(lowest_level == TRUE) -> dat

dat_bls_1 <- bls_api(dat$series_id[1:50], 
                   startyear = 2006, 
                   endyear = 2021, 
                   Sys.getenv("BLS_KEY"))

dat_bls_2 <- bls_api(dat$series_id[51:100], 
                   startyear = 2006, 
                   endyear = 2021, 
                   Sys.getenv("BLS_KEY"))

dat_bls_3 <- bls_api(dat$series_id[101:150], 
                   startyear = 2006, 
                   endyear = 2021, 
                   Sys.getenv("BLS_KEY"))

dat_bls_4 <- bls_api(dat$series_id[151:180], 
                   startyear = 2006, 
                   endyear = 2021, 
                   Sys.getenv("BLS_KEY"))


rbind(dat_bls_1,
      dat_bls_2,
      dat_bls_3,
      dat_bls_4) -> dat_all

key <- data.frame(name = raw_dat$name, code = raw_dat$series_id)
cat_key <- data.frame(key = raw_dat$name, category = raw_dat$category)



dat_all %>% 
  mutate(date = paste0(periodName, " ", year, " 01"),
         date = as.Date(date, "%B %Y %d")) %>% 
  rename(code = seriesID) %>% 
  left_join(key, by = "code") %>% 
  select(date, name, value) %>% 
  pivot_wider(names_from = "name", values_from = "value") %>% 
  arrange(date) -> dat_wide



dat_wide[,-1] %>%
  map_df(.f = function (x) {(x / lag(x) - 1) * 100}) %>% 
  mutate(mean = rowMeans(., na.rm = TRUE),
         sd = rowSds(as.matrix(.), na.rm = TRUE),
         mean_plussd = mean + sd,
         mean_lesssd = mean - sd,
         x_inter = 0) -> dat_wide_mm 

dat_wide[,-1] %>%
  map_df(.f = function (x) {(x / lag(x, 12) - 1) * 100}) %>% 
  mutate(mean = rowMeans(., na.rm = TRUE),
         sd = rowSds(as.matrix(.), na.rm = TRUE),
         mean_plussd = mean + sd,
         mean_lesssd = mean - sd,
         x_inter = 0) -> dat_wide_yy

dats <- list()
cbind(dat_wide[,1], dat_wide_mm) -> dats[[1]]
cbind(dat_wide[,1], dat_wide_yy) -> dats[[2]]


dats_final <- list()
for (i in 1:2) {
 dats[[i]] %>% 
  pivot_longer(-date, names_to = "key", values_to = "values") %>% 
  filter(complete.cases(.),
         date > "2019-05-01") %>% 
  left_join(cat_key, by = "key") -> dats_final[[i]]
}


ctitle_1 <- "Distribution of CPI component year-over-year changes"
unit_label <- "Number of components within 0.8%pt bins"
x_axis_lab <- "Year-over-year change in component index level, %"
source_1 <- "Source: BLS, @benbakkum."

dats_final[[2]] %>% 
  filter(!key %in% c("mean", "sd", "mean_plussd", "mean_lesssd", "x_inter")) %>% 
  ggplot(aes(x = values, fill = category, label = format(date, "%b '%y"))) +
  geom_histogram(binwidth = 0.8) +
  geom_vline(data = dats_final[[2]] %>% filter(key == "x_inter"),
             aes(xintercept = values, frame = date), 
             linetype = "dashed") +
  geom_vline(data = dats_final[[2]] %>% filter(key == "mean"), 
             aes(xintercept = values, frame = date, color = "Mean"), 
             linetype = "dashed",
             size = 1) +
  geom_vline(data = dats_final[[2]] %>% filter(key == "mean_plussd"), 
             aes(xintercept = values, frame = date, color = "Mean +1sd"), 
             linetype = "dotted",
             size = 1) +
  geom_vline(data = dats_final[[2]] %>% filter(key == "mean_lesssd"), 
             aes(xintercept = values, frame = date, color = "Mean -1sd"), 
             linetype = "dotted",
             size = 1) +
  geom_text(data = dats_final[[2]] %>% filter(key == "Bread"),
            aes(-13, 26), 
            size = 13, 
            alpha = .8, 
            color = cpal("border grey"),
            family = "Open Sans") +
  scale_color_manual(values = c(cpal("red"),  cpal("border grey"), cpal("border grey"))) +
  scale_x_continuous(expand = c(0,0), limits = c(-20, 20)) +
  scale_y_continuous(expand = c(0,0), breaks = seq(0, 35, 5)) +
  scale_fill_manual(values = c(cpal("paleblue"), cpal("paleorange"), cpal("green"), 
                               cpal("blue"), cpal("red"))) +
  my_theme +
  theme(axis.title.x = element_text(size = 14, color = cpal("text"), family = "Open Sans"),
        legend.margin = margin(t = 0, unit = "cm")) +
  guides(color = guide_legend(nrow = 2, 
                              override.aes = list(linetype = c("dashed", "dotted", "dotted"))),
         fill = guide_legend(nrow = 2)) +
  labs(title = ctitle_1, subtitle = unit_label, caption = source_1) +
  xlab(label = x_axis_lab) +
  transition_states(date, transition_length = 2, wrap = FALSE) +
  ease_aes("linear") -> chart1



anim1 <- animate(chart1, renderer = gifski_renderer(loop = T), nframe = 260,
                 width = 615, height = 470, type = "cairo", end_pause = 50)

anim1
```

Looking at their contributions to the overall rate of year-over-year inflation provides a cleaner look at the relative trends of goods and services prices. Based on this perspective, core goods appear in large responsible for the updraft in inflation that everyone's talking about, though core services' contribution has steadily increased over the last four months.

```{r echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE}

xlist <- list.files(path = "D:\\Documents\\R directory\\inflation\\relative_importance_standard", pattern = "*.csv")

csvs <- list()

for(i in xlist) { 
  csvs[[i]] <- read.csv(paste0("D:\\Documents\\R directory\\inflation\\relative_importance_standard\\", i), check.names = F)
}


do.call(rbind, csvs) %>% 
  as_tibble(rownames = NA) %>% 
  rownames_to_column() %>% 
  rename(file = rowname) %>% 
  mutate(file = substr(file, 1, 8)) %>% 
  separate(file, c("month", "year"), "_") %>% 
  mutate(month = case_when(month == "jan" ~ "01",
                           month == "feb" ~ "02",
                           month == "mar" ~ "03",
                           month == "apr" ~ "04",
                           month == "may" ~ "05",
                           month == "jun" ~ "06",
                           month == "jul" ~ "07",
                           month == "aug" ~ "08",
                           month == "sep" ~ "09",
                           month == "oct" ~ "10",
                           month == "nov" ~ "11",
                           month == "dec" ~ "12")) %>% 
  mutate(date = paste0(year, "-", month, "-01"),
         date = as.Date(date, "%Y-%m-%d"),
         Component = gsub("\\s*\\([^\\)]+\\)","", Component)) %>% 
  select(-month, -year) %>% 
  filter(Component %in% c("Commodities less food and energy commodities",
                          "Services less energy services")) %>% 
  pivot_wider(names_from = "Component", values_from = "RI") %>% 
  set_names("date", "core_goods", "core_services") %>% 
  arrange(date) %>% 
  mutate(total = core_goods + core_services,
         core_goods_wt = core_goods / total,
         core_services_wt = core_services / total) %>% 
  select(-total, -core_goods, -core_services) -> dat_wts


raw_dat <- getSymbols(c("CUSR0000SACL1E", "CUSR0000SASLE", "CPILFESL"), src='FRED', auto.assign=T) 

key <- c("core_goods", "core_services", "core")


merge(CUSR0000SACL1E, CUSR0000SASLE, CPILFESL)  %>% 
  as_tibble(rownames = NA) %>% 
  rownames_to_column() %>% 
  set_names("date", key) %>% 
  mutate(date = as.Date(date, "%Y-%m-%d"),
         core_goods = (core_goods / lag(core_goods,12) - 1) * 100,
         core_services = (core_services / lag(core_services,12) -1) * 100,
         core = (core / lag(core,12) -1) * 100) %>% 
  filter(date > "2016-12-31") -> dat_yy
  


dat_wts %>% 
  right_join(dat_yy, by = "date") %>% 
  mutate(core_goods_cont = core_goods * lag(core_goods_wt, 12),
         core_services_cont = core_services * lag(core_services_wt, 12),
         cont_total = core_goods_cont + core_services_cont) %>% 
  select(date, core_goods_cont, core_services_cont, core) %>% 
  filter(date > "2017-12-31") %>% 
  pivot_longer(-date, names_to = "key", values_to = "values") %>% 
  mutate(key = case_when(key == "core_goods_cont" ~ "Core goods contribution",
                         key == "core_services_cont" ~ "Core services contribution",
                         key == "core" ~ "Core CPI"),
         placeholder = "x") -> dat_long


ctitle_1 <- "Contributions of goods and services to core CPI"
unit_label <- "%, year-over-year change in core CPI"
source_1 <- "Source: BLS, FRED, @benbakkum."



dat_long %>% 
  filter(!key == "Core CPI") %>% 
  ggplot(aes(date, values, fill = key)) +
  geom_col() +
  geom_line(data = dat_long %>% filter(key == "Core CPI"),
            aes(date, values, color = key),
            show.legend = F,
            size = 1.2) +
  scale_fill_manual(values = c(cpal("red"), cpal("paleorange"), cpal("blue"))) +
  scale_color_manual(values = cpal("red")) +
  scale_x_date(expand = c(0,0)) +
  my_theme +
  labs(title = ctitle_1, subtitle = unit_label, caption = source_1) +
  guides(fill = guide_legend(override.aes = list(size=9),
                             nrow = 3)) +
  transition_manual(placeholder) -> p
  
anim1 <- animate(p, renderer = gifski_renderer(loop = T), nframe = 1,
                 width = 615, height = 460, type = "cairo")

anim1

```

As can be seen in the chart below, the current jump in core goods prices flies in the face of decades of flatlining prices in the category. Structural forces of productivity growth, competition from abroad, and other factors have surely kept a lid on core goods prices this century, and I expect once the pandemic's disruptions abate that those forces will reassert themselves.

```{r echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE}
raw_dat <- getSymbols(c("CUSR0000SACL1E", "CUSR0000SASLE"), src='FRED', auto.assign=T) 

key <- c("core_goods", "core_services")


merge(CUSR0000SACL1E, CUSR0000SASLE)  %>% 
  as_tibble(rownames = NA) %>% 
  rownames_to_column() %>% 
  set_names("date", key) %>% 
  mutate(date = as.Date(date, "%Y-%m-%d")) %>% 
  filter(date > "1999-12-31") %>% 
  mutate(core_goods = (core_goods / core_goods[1]) * 100,
         core_services = (core_services / core_services[1]) * 100) %>% 
  pivot_longer(-date, names_to = "key", values_to = "values") %>% 
  mutate(key = case_when(key == "core_goods" ~ "Core goods",
                         key == "core_services" ~ "Core services")) -> dat


ctitle_1 <- "Core goods and core services consumer price indices"
unit_label <- "1/1/2000 = 100"
source_1 <- "Source: BLS, FRED, @benbakkum."


dat %>% 
  ggplot(aes(date, values, color = key)) +
  geom_line(size = 1.4, show.legend = F) +
  geom_point(size = 3.5) +
  scale_color_manual(values = c(cpal("orange"), cpal("blue"))) +
  geom_text(aes(date, values, label = round(values, 0)), 
                fontface = "bold",
                hjust = -0.5,
            size = 5,
            show.legend = F) +
  scale_x_date(expand = c(0,0),
               limits = c(as.Date("2000-01-01"), as.Date(tail(dat,1)$date %m+% months(24)))) +
  my_theme +
  guides(colour = guide_legend(override.aes = list(size=9))) +
  labs(title = ctitle_1, subtitle = unit_label, caption = source_1) +
  transition_reveal(date, keep_last = TRUE) +
  view_follow(fixed_x = TRUE, fixed_y = FALSE) -> chart1




anim1 <- animate(chart1, renderer = gifski_renderer(loop = T), nframe = 280,
                 width = 615, height = 450, type = "cairo", end_pause = 200)

anim1
```

Another way to breakdown the year-over-year series of core CPI involves ascribing the overall change in its current level from the level 12 months ago to different periods in-between. The bulk of what has pushed the year-over-year inflation so high came in the last nine months, so this much of this increase won't begin to roll off for another several months even if inflation moderates from here. 

```{r echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE}
raw_dat <- getSymbols(c("CPILFESL"), src='FRED', auto.assign=T) 

key <- c("infl")



ctitle_1 <- "Contributions by timing of price increases over last 12 months"
unit_label <- "%, year-over-year change in core CPI"
source_1 <- "Source: BLS, FRED, @benbakkum."



CPILFESL %>% 
  as_tibble(rownames = NA) %>% 
  rownames_to_column() %>% 
  set_names("date", key) %>% 
  mutate(date = as.Date(date, "%Y-%m-%d")) %>% 
  filter(date > "2017-12-31") %>% 
  mutate(yy = (infl / lag(infl, 12) - 1) * 100,
         ten_twelve = (lag(infl, 9) / lag(infl, 12) - 1) * 100,
         seven_nine = ((lag(infl, 6) / lag(infl, 12) - 1) * 100) - ten_twelve,
         four_six = ((lag(infl, 3) / lag(infl, 12) - 1) * 100) - seven_nine - ten_twelve,
         one_three = yy - four_six - seven_nine - ten_twelve,
         total = one_three + four_six + seven_nine + ten_twelve) %>% 
  select(-infl, -total, - yy) %>% 
  pivot_longer(-date, names_to = "key", values_to = "values") %>% 
  mutate(key = case_when(key == "ten_twelve" ~ "10-12 months ago",
                         key == "seven_nine" ~ "7-9 months ago",
                         key == "four_six" ~ "4-6 months ago",
                         key == "one_three" ~ "1-3 months ago"),
         key = factor(key, levels = c("10-12 months ago", "7-9 months ago", 
                                      "4-6 months ago", "1-3 months ago")),
         placeholder = "x") %>% 
  ggplot(aes(date, values, fill = key)) +
  geom_col() +
  geom_hline(yintercept = 0, linetype = "dashed", color = cpal("border grey"), size = 1.1) +
  scale_x_date(expand = c(0,0),
               breaks = "3 months",
               date_labels = "%b '%y") +
  scale_fill_manual(values = c(cpal("light grey"), cpal("paleorange"),  cpal("paleblue"), cpal("blue"))) +
  labs(title = ctitle_1, subtitle = unit_label, caption = source_1) +
  my_theme +
  theme(axis.text.x = element_text(angle = 60, vjust = 0.6)) +
  guides(fill = guide_legend(override.aes = list(size=9),
                             nrow = 2)) +
  transition_manual(placeholder) -> p




anim1 <- animate(p, renderer = gifski_renderer(loop = T), nframe = 1,
                 width = 615, height = 460, type = "cairo")

anim1


```

Hypothetically if inflation begins to moderate from here — though I'm cognizant of upside risks, especially in the form of housing prices — year-over-year core CPI would remain elevated for the next several months before beginning to decline. The chart below extrapolates year-over-year inflation with the following months each exhibiting the average monthly rate of inflation of the last five years. Note that the Fed prefers to base its policy on the PCE price index, which would likely be lower based on the smaller weight of housing in that index. 

```{r echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE}
dat <- bls_api(c("CUSR0000SA0", "CUSR0000SA0L1E"), 
              startyear = 2006, 
              endyear = 2021, 
              Sys.getenv("BLS_KEY"))

data.frame(name = c("headline_cpi", "core_cpi"),
           code = c("CUSR0000SA0", "CUSR0000SA0L1E")) -> key

rolling_mean <- rollify(mean, window = 60)

dat %>% 
  mutate(date = paste0(periodName, " ", year, " 01"),
         date = as.Date(date, "%B %Y %d")) %>% 
  rename(code = seriesID) %>% 
  left_join(key, by = "code") %>% 
  select(value, date, name) %>% 
  pivot_wider(names_from = "name", values_from = "value") %>%
  arrange(date) %>% 
  mutate(headline_mm      =  (headline_cpi / lag(headline_cpi, 1) - 1) * 100,
         core_mm          =  (core_cpi / lag(core_cpi, 1) - 1) * 100,
         headline_mm_avg  =   rolling_mean(headline_mm),
         core_mm_avg      =   rolling_mean(core_mm),
         headline_dev     = -(headline_mm - headline_mm_avg),
         core_dev         = -(core_mm - core_mm_avg),
         headline_proj    =   headline_cpi,
         core_proj        =   core_cpi) -> df




x <- nrow(df)
next_month <- ymd(pull(df[,1])) %m+% months(1)
for (i in 1:18) {
  df[x + i,] <- NA
  
  df[x + i,1] <- ymd(pull(df[x + i -1,1])) %m+% months(1)
  
  df$headline_proj[x + i] <- (1 + df$headline_mm_avg[x]/100) * df$headline_proj[x + i - 1]
  
  df$core_proj[x + i] <- (1 + df$core_mm_avg[x]/100) * df$core_proj[x + i - 1]
  
  df$headline_mm_avg[x + i] <- df$headline_mm_avg[x]
  
  df$core_mm_avg[x + i] <- df$core_mm_avg[x]
}

df %>% 
  mutate(headline_dev_lag =   lag(headline_dev,12),
         core_def_lag     =   lag(core_dev,12),
         headline_adj     =   (1 + lag(headline_mm_avg,1)/100) * lag(headline_proj,1),
         core_adj         =   (1 + lag(core_mm_avg,1)/100) * lag(core_proj,1),
         headline_yy      =   ((headline_proj / lag(headline_proj,12) - 1) * 100),
         core_yy          =   ((core_proj / lag(core_proj,12) - 1) * 100),
         headline_yy_adj  =   ((headline_proj / lag(headline_adj,12) - 1) * 100),
         core_yy_adj      =   ((core_proj / lag(core_adj,12) - 1) * 100),
         headline_be      =   headline_yy - headline_yy_adj,
         core_be          =   core_yy - core_yy_adj,
         headline_yy_ex   =   case_when(date < ymd(next_month) ~ NA_real_,
                                        date >= ymd(next_month) ~ headline_yy),
         core_yy_ex       =   case_when(date < ymd(next_month) ~ NA_real_,
                                        date >= ymd(next_month) ~ core_yy),
         headline_yy_act  =   case_when(date >= ymd(next_month) ~ NA_real_,
                                        date < ymd(next_month) ~ headline_yy),
         core_yy_act      =   case_when(date >= ymd(next_month) ~ NA_real_,
                                        date < ymd(next_month) ~ core_yy)) -> df_all 

eight_mths_ago <- ymd(pull(df[x,1])) %m+% months(-8)
five_mths_later <- ymd(pull(df[x,1])) %m+% months(5)
curr_mth <- ymd(pull(df[x,1])) + days(14)
twenty_mths_ago <- ymd(pull(df[x,1])) %m+% months(-20)
sixteen_mths_later <- ymd(pull(df[x,1])) %m+% months(16)

df_all %>% 
  select(date, headline_yy_ex, core_yy_ex) %>% 
  pivot_longer(-date, names_to = "key", values_to = "values") %>%
  mutate(key = case_when(key == "core_yy_ex" ~ "Extrapolated core CPI",
                         key == "headline_yy_ex" ~ "Extrapolated headline CPI"),
         placeholder = "X") %>% 
  filter(date > twenty_mths_ago & date < sixteen_mths_later) -> ex_cht_dat

df_all %>% 
  select(date, headline_yy_act, core_yy_act) %>% 
  pivot_longer(-date, names_to = "key", values_to = "values") %>%
  mutate(key = case_when(key == "core_yy_act" ~ "Core CPI",
                         key == "headline_yy_act" ~ "Headline CPI"),
         placeholder = "X") %>% 
  filter(date > twenty_mths_ago & date < sixteen_mths_later) -> act_cht_dat

ex_cht_dat$key <- factor(ex_cht_dat$key, c("Extrapolated core CPI", "Extrapolated headline CPI"))
act_cht_dat$key <- factor(act_cht_dat$key, c("Core CPI", "Headline CPI"))
        
ctitle_1 <- "Year-over-year CPI extrapolated w/ avg monthly change"
unit_label <- "%, y/y"
source_1 <- "Source: BLS, @benbakkum."

ggplot() +
  geom_step(data = ex_cht_dat, aes(date, values, color = key, linetype = key), size = 1) +
  geom_step(data = act_cht_dat, aes(date, values, color = key, linetype = key), size = 1) +
  scale_color_manual(values = c(cpal("blue"), cpal("blue"), cpal("green"), cpal("green"))) +
  scale_linetype_manual(values = c("solid", "dashed", "dashed", "solid")) +
  scale_y_continuous(breaks = seq(0,5.5, by = 0.5)) +
  scale_x_date(expand = c(0,0), date_breaks = "6 months", date_labels = "%b '%y") +
  geom_vline(xintercept = ymd(curr_mth), linetype = "dashed", color = cpal("border grey"), size = 1) +
  labs(title = ctitle_1, subtitle = unit_label, caption = source_1) +
  my_theme +
  guides(color = guide_legend(ncol = 2, byrow = T,
                              override.aes = list(linetype = c("solid", "twodash", "twodash", "solid"),
                                                  size = 1))) +
  transition_manual(placeholder) -> p

anim1 <- animate(p, renderer = gifski_renderer(loop = T), nframe = 1,
                 width = 615, height = 420, type = "cairo")

anim1

```

For reference, various rates of inflation for important components of CPI can be found in the table below.

```{r inflation_table, echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE}
getSymbols(c("CPIAUCSL", "CPILFESL", "CPIFABSL", "CPIENGSL",
          "CUSR0000SACL1E", "CUSR0000SASLE", 
          "CPIAPPSL", "CUSR0000SETA02",
          "CUSR0000SETG01", "CUSR0000SAH1", "CUSR0000SAM2", 
          "CUSR0000SAE1", "CUSR0000SAS4"),
        src='FRED', auto.assign=T) -> fred_import

c("Headline CPI", "Core CPI (~79)", "Food (~14)", "Energy (~7)",
  "Core goods (~21)", "Core services (~58)", 
  "Apparel (~3)", "Used cars and trucks (~3)",
  "Airline fares (~1)", "Shelter (~33)", "Medical care services (~7)",
  "Education (~3)", "Transportation services (~5)") -> key

merge(CPIAUCSL, CPILFESL, CPIFABSL, CPIENGSL,
      CUSR0000SACL1E, CUSR0000SASLE, 
      CPIAPPSL, CUSR0000SETA02,
      CUSR0000SETG01, CUSR0000SAH1, CUSR0000SAM2, 
      CUSR0000SAE1, CUSR0000SAS4) %>% 
  as_tibble(rownames = NA) %>% 
  rownames_to_column() %>% 
  set_names("date", key) %>%
  filter(date > "2018-12-31") -> dat

dat %>% 
  select(-date) %>% 
  map_df(.f = function (x) {round((x / lag(x) - 1) * 100,1)}) %>%
  tail(8) %>% 
  t() %>% 
  as_tibble(rownames = NA) %>% 
  rownames_to_column() %>% 
  set_names("Index", format(as.Date(tail(dat,8)$date), "%b-%y")) %>% 
  mutate(Period = "m/m, %",
         r = row_number(),
         type = 1) -> dat_mm
  
dat %>% 
  select(-date) %>% 
  map_df(.f = function (x) {round(((x / lag(x,3))^4 - 1) * 100,1)}) %>%
  tail(8) %>% 
  t() %>% 
  as_tibble(rownames = NA) %>% 
  rownames_to_column() %>% 
  set_names("Index", format(as.Date(tail(dat,8)$date), "%b-%y")) %>% 
  mutate(Period = "3m ann, %",
         r = row_number(),
         type = 2) -> dat_3m

dat %>% 
  select(-date) %>% 
  map_df(.f = function (x) {round((x / lag(x,12) - 1) * 100,1)}) %>%
  tail(8) %>% 
  t() %>% 
  as_tibble(rownames = NA) %>% 
  rownames_to_column() %>% 
  set_names("Index", format(as.Date(tail(dat,8)$date), "%b-%y")) %>% 
  mutate(Period = "y/y, %",
         r = row_number(),
         type = 3) -> dat_yy
  
rbind(dat_mm, dat_3m, dat_yy) %>% 
  arrange(r, type) %>% 
  select(-r, -type) %>% 
  select(Index, Period, everything()) %>% 
  as.data.frame() -> df_tbl


df_tbl %>% 
  rename(" " = Period) %>% 
  mutate(m1_chg = .[[4]] - .[[3]],
         m2_chg = .[[5]] - .[[4]],
         m3_chg = .[[6]] - .[[5]],
         m4_chg = .[[7]] - .[[6]],
         m5_chg = .[[8]] - .[[7]],
         m6_chg = .[[9]] - .[[8]],
         m7_chg = .[[10]] - .[[9]]) -> df_chgs

df_chgs[,c(1,2,11,4,12,5,13,6,14,7,15,8,16,9,17,10)] -> df_tbl


## decline is cooling, increasing is heating up
dec_is_good <- function(i, j) {
  
  y <- as.numeric(df_tbl[j,i-1])
  
  case_when(y <  -1              ~ color_tile("#90d1fc", "#90d1fc")(df_tbl[j,i]),
           (y >= -1) & (y < -.3)   ~ color_tile("#d1ecff", "#d1ecff")(df_tbl[j,i]),
           (y >= -.3) & (y <= .3)           ~ as.character(df_tbl[j,i]),
           (y <=  1) & (y > .3)   ~ color_tile("#ffd6d6", "#ffd6d6")(df_tbl[j,i]),
            y >   1              ~ color_tile("#f59393", "#f59393")(df_tbl[j,i])) 
}


for (j in c(1:39)) {
  for (i in seq(4,(ncol(df_tbl)),2)) {
    df_tbl[j,i] <- dec_is_good(i, j)
  }
}


## https://github.com/haozhu233/kableExtra/issues/217
df_tbl %>% 
  kable(escape = F,
        align = c('l', 'l', rep('r', 14))) %>% 
  kable_paper("hover", full_width = F, html_font = "Open Sans") %>%
  remove_column(3) %>% 
  remove_column(4) %>% 
  remove_column(5) %>% 
  remove_column(6) %>% 
  remove_column(7) %>% 
  remove_column(8) %>% 
  remove_column(9) %>% 
  remove_column(1) %>% 
  group_rows("Headline CPI", 1, 3) %>% 
  group_rows("Core CPI (~79)", 4, 6) %>% 
  group_rows("Food (~14)", 7, 9) %>% 
  group_rows("Energy (~7)", 10, 12) %>% 
  group_rows("Core goods (~21)", 13, 15) %>% 
  group_rows("Core services (~58)", 16, 18) %>% 
  group_rows("Apparel (~3)", 19, 21) %>% 
  group_rows("Used cars and trucks (~3)", 22, 24) %>% 
  group_rows("Airline fares (~1)", 25, 27) %>% 
  group_rows("Shelter (~33)", 28, 30) %>% 
  group_rows("Medical care services (~7%)", 31, 33) %>% 
  group_rows("Education (~3%)", 34, 36) %>% 
  group_rows("Transportation services (~5%)", 37, 39) %>% 
  add_footnote("Number next to component name represents its weight in the overall index. \nColoring represents the degree of change in the rate of inflation from the \nmonth before. Source: BLS, @benbakkum.", notation = "none") -> table



table
```

