---
title: "COVID-19 -> Europe's Resurgence"
date: 2020-08-30T21:13:14-05:00
draft: false
---
&nbsp;
&nbsp;

In a previous post from June called [The American Surge](https://www.macrochronicles.com/blog/the-american-surge/), I wrote about the debate at the time over whether the US had entered a second wave given it seemed clear that the first wave to hit the country had never really passed. I also suggested that fire may be a more fitting metaphor and that "in the places where the fire has been brought down to a smoldering remnant, a handful of embers can still exist that erupt again when oxygen returns." Whether you call it a fresh flame or another wave, unfortunately it's happening in Europe right now. After lockdowns successfully dampened the spread of COVID-19 across the continent through most of the spring and summer, many major European countries are now again reporting new cases in the thousands each day.    

```{r setup-new_cases, include=FALSE}
## data from https://data.humdata.org/dataset/novel-coronavirus-2019-ncov-cases

library(tidyverse)
library(xts)
library(tibbletime)
library(gganimate)
library(gifski)
library(Cairo)
library(magick)
library(scales)
library(ggrepel)

raw_dat <- read_csv("D:\\Documents\\R directory\\covid19\\cases.csv")

dat <- raw_dat %>%
  as_tibble() %>% 
  rename(Province = `Province/State`, Country = `Country/Region`) %>% 
  select(-Province,-Lat,-Long) %>% 
  group_by(Country) %>%
  summarise_all(sum) %>%
  rownames_to_column() %>% 
  gather(var, value, -rowname) %>% 
  spread(rowname, value) %>% 
  mutate(var = as.Date(var, "%m/%d/%Y")) %>% 
  arrange(var)


colnames(dat) <- dat[nrow(dat),] 
  
rec_count <- tail(dat[-nrow(dat),],1)
my_names <- colnames(dat)[as.numeric(rec_count[1,]) > 1000][-1]
dat_sm <- dat %>%
  rename(date = 'NA') %>% 
  select(date, my_names, Singapore)

dat_sm2 <- dat_sm[-nrow(dat),] %>% 
  rename('S Korea' = 'Korea, South') %>%
  rename(UK = 'United Kingdom') %>% 
  mutate(date = as.Date(date, "%m/%d/%Y")) %>%
  arrange(date)

dat_sm3 <- dat_sm2[,-1] %>% 
  map(.f = function(x) {as.numeric(x)}) %>% 
  cbind(dat_sm2[,1]) %>% 
  select(date, everything())

rolling_mean <- rollify(mean, window = 7)

dat_sm3.5 <- dat_sm3[,-1] %>% 
  map(.f = function (x) {x - lag(x, 1)}) %>%
  map(.f = function (x) {rolling_mean(x)}) %>%
  as_data_frame() %>% 
  cbind(dat_sm3$date) %>% 
  rename(date = 'dat_sm3$date') %>% 
  select(date, everything()) %>% 
  as_data_frame()
  
dat_sm4 <- dat_sm3.5 %>% 
  gather(key, value, -date)

keep_cntry <- dat_sm4 %>% 
  na.omit() %>% 
  group_by(key) %>% 
  summarise(max = max(value)) %>% 
  filter(max > 2500) %>% 
  select(-max)

dat_sm5 <- dat_sm4 %>% 
  filter(key %in% keep_cntry$key) %>% 
  arrange(key, date) %>% 
  mutate(crossover = case_when((value >= 20 & lag(value,1) < 20) | 
                               (key != lag(key,1))~ 1,
                               TRUE ~ 0))
  

dat_sm5[dat_sm5$date == "2020-06-14" & dat_sm5$key == "China", "crossover"] <- 0
dat_sm5[dat_sm5$date == "2020-07-01" & dat_sm5$key == "China", "crossover"] <- 0
dat_sm5[dat_sm5$date == "2020-07-09" & dat_sm5$key == "China", "crossover"] <- 0
dat_sm5[dat_sm5$date == "2020-05-06" & dat_sm5$key == "Kyrgyzstan", "crossover"] <- 0	
dat_sm5[dat_sm5$date == "2020-05-18" & dat_sm5$key == "Kyrgyzstan", "crossover"] <- 0


dat_sm6 <- dat_sm5 %>%
  group_by(key, idx = cumsum(crossover == 1L)) %>% 
  mutate(counter = row_number(),
         counter = as.numeric(counter),
         phase = case_when((idx %% 2) == 0 ~ "off",
                           (idx %% 2) != 0 ~ "on")) %>%
  ungroup() %>% 
  mutate(days = case_when(phase == "off" ~ 0,
                          phase == "on" ~ counter)) %>% 
  select(-crossover, -idx, -counter, -phase) %>% 
  filter(days > 0) %>% 
  mutate(days = days - 1)

dat_c <- dat_sm6 
dat_lab <- dat_c %>% filter(key == "Argentina" | key == "Brazil" | key == "China" |
                            key == "France" | key == "India" | key == "Iran" |
                            key == "Italy" | key == "Peru" | 
                            key == "Spain" |
                            key == "US")

#write.csv(dat_sm5, "D:\\Documents\\R directory\\covid19\\new_cases.csv")

```


```{r chart-new_cases, echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE}
fmt_dcmls <- function(decimals = 0) {
  function(x) as.character(format(round(as.numeric(x), 0), nsmall=0, big.mark=","))
}

colors <- c('paleblue' = "#a6cee3", 'blue' = "#1f78b4", 'green' = "#33a02c", 'palegreen' = "#b2df8a",
            'red' = "#e31a1c", 'palered' = "#fb9a99", 'orange' = "#ff7f00", 'paleorange' = "#fdbf6f",
            'purple' = "#6a3d9a", 'palepurple' = "#cab2d6", 'border grey' = "#808080", 'brown' = '#b15928',
            'text' = "#222222" , 'light grey' = "#CDCDCD")

cpal <- function(...) {
  cols <- c(...)
  if (is.null(cols))
    return(colors)
  colors[[cols]]
}

ctitle_1 <- "COVID-19 daily new cases"
unit_label <- "Daily change in cases, 7 day moving average"
source_1 <- "Source: JHU, WHO, @benbakkum." 

line_cols <- c("Argentina" = cpal("green"),
               "Bangladesh" = "#CDCDCD",
               "Brazil" = cpal("green"),
               "Chile" = "#CDCDCD",
               "China" = cpal("blue"),
               "Colombia" = "#CDCDCD",
               "France" = cpal("purple"),
               "Germany" = "#CDCDCD",
               "India" = cpal("blue"),
               "Iran" = cpal("blue"),
               "Iraq" = "#CDCDCD",
               "Italy" = cpal("purple"),
               "Kazakhstan" = "#CDCDCD",
               "Kyrgyzstan" = "#CDCDCD",
               "Mexico" = "#CDCDCD",
               "Pakistan" = "#CDCDCD",
               "Peru" = cpal("green"),
               "Philippines" = "#CDCDCD",
               "Russia" = "#CDCDCD",
               "Saudi Arabia" = "#CDCDCD",
               "South Africa" = "#CDCDCD",
               "Spain" = cpal("purple"),
               "Turkey" = "#CDCDCD",
               "UK" = "#CDCDCD",
               "US" = cpal("brown"))
               


chart1 <- ggplot(data = dat_c, aes(x = days, y = value, color = key, label = format(date, '%b %d'))) +
  geom_text(data = filter(dat_c, key == "China"), aes(125, 40), color = "#808080", size = 13, alpha = .8, 
            family = "Open Sans") +
  geom_line(size = 1.1, alpha = 0.65) +
  scale_y_log10(labels = fmt_dcmls(0), 
                breaks = c(0, 1, 2, 5, 10, 20, 50, 100, 200, 500, 1000, 2000, 5000,
                                                  10000, 20000, 50000),
                limits = c(10, 110000),
                expand = c(0,0)) +
  scale_color_manual(values = line_cols) +
  geom_point(data = dat_lab, aes(x = days, y = value), size = 2.7) +
  geom_text_repel(data = dat_lab, aes(x = days, y = value, label = key), nudge_x = 3.5, nudge_y = 0.15,
                                show.legend = FALSE, size = 4, family = "Open Sans", fontface = "bold") +
  scale_x_continuous(limits = c(0, 200), expand = c(0,0)) +
  theme(
    legend.position = "none",
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = cpal("light grey"), size = 0.21),
    panel.border = element_rect(fill = NA, color = cpal("border grey"), size = 0.3),
    plot.title = element_text(size = 21, color = "black", family = "Open Sans Light"),
    plot.subtitle = element_text(size = 14, color = cpal("text"), family = "Open Sans"),
    plot.caption = element_text(size = 12.5, color = cpal("text"), family = "Open Sans"),
    text = element_text(size = 15, color = cpal("text"), family = "Open Sans"),
    legend.text = element_text(size = 15, color = cpal("text"), family = "Open Sans"),
    axis.text.x = element_text(size = 14, color = cpal("text"), family = "Open Sans",
                               margin = unit(c(0.35, 0.25, 0.25, 0.25), "cm")),
    axis.text.y = element_text(size = 14, color = cpal("text"), family = "Open Sans",
                               margin = unit(c(0.35, 0.25, 0.25, 0.25), "cm")),
    axis.title.y = element_blank(),
    axis.title.x = element_text(size = 14, color = cpal("text"), family = "Open Sans",
                               margin = unit(c(0.05, 0.05, 0.05, 0.05), "cm")),
    axis.ticks.length = unit(-0.15, "cm"),
    axis.ticks = element_line(color = cpal("border grey"), size = 0.3),
    plot.margin = unit(c(0.3, 0.5, 0.3, 0.05), "cm")
  ) + 
  labs(title = ctitle_1, subtitle = unit_label, x = "Number of days since averaging over 20 new cases a day",
       caption = paste("Data as of ", format(tail(dat_c$date,1), "%b %d, %Y"), ". ", source_1[1], sep = "")) +
  transition_reveal(date, keep_last = TRUE) +
  ease_aes('linear')
  
anim1 <- animate(chart1, renderer = gifski_renderer(loop = T), nframe = 475,
                 width = 615, height = 420, type = "cairo", end_pause = 200)  

anim1
```

The cold weather that will come with fall, drawing people indoors, and the seeming lack of political will to enforce lockdowns with the type of severity of those in March and April, does not augur well for the outbreaks' containment. French President Emmanuel Macron recently expressed a preference for "very localized strategies" rather than a national lockdown. In Spain, 17 regional governments have taken over the response to the pandemic after the end of the state of emergency in June, creating an unorganized patchwork of different rules that will vary in effectiveness. The director of Spain's health emergency center says that "nobody should be in any doubt, things are not going well."^[Marc Santora, ["With Coronavirus Cases Surging, Europe Braces for New Phase in Pandemic"](https://www.nytimes.com/2020/08/21/world/europe/coronavirus-second-wave.html), _The New York Times_, August 21, 2020.] 

In the US, things have stopped getting worse, but they aren't getting much better. As a positive sign, the percentage of conducted tests that are positive for the virus continues to edge lower. The number of tests being administered each day, however, has also ebbed.

```{r tests, echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE}

### https://covidtracking.com/api/v1/us/daily.csv 

colors <- c('aqua' = "#45BCC9", 'green' = "#648C2E", 'deep blue' = "#00567D",
            'espresso' = "#310008", 'sand' = "#E1D1A7", 'orange' = "#F38B00",
            'red' = "#A51D31", 'dark grey' = "8E8279", 'border grey' = "#808080",
            'text' = "#222222", 'light grey' = "#CDCDCD")

cpal <- function(...) {
  cols <- c(...)
  if (is.null(cols))
    return(colors)
  colors[[cols]]
}

raw_dat <- read_csv("D:\\Documents\\R directory\\covid19\\nat_daily.csv")

dat <- raw_dat %>% 
  as_data_frame() %>%
  select(date, positiveIncrease, negativeIncrease, totalTestResultsIncrease) %>% 
  mutate(date = as.Date(as.character(date), "%Y%m%d"),
         percPositive = (positiveIncrease / (positiveIncrease + negativeIncrease)) * 100) %>% 
  rename(tests = totalTestResultsIncrease) %>% 
  select(date, percPositive, tests)

ctitle_1 <- "Tests conducted per day"
source_1 <- "The COVID Tracking Project, @benbakkum." 


fmt_dcmls <- function(decimals = 0) {
  function(x) as.character(format(round(as.numeric(x), 0), nsmall=0, big.mark=","))
}

chart1 <- dat %>% 
  filter(date > "2020-04-30") %>% 
  ggplot(aes(x = date, y = tests)) +
  geom_bar(stat = 'identity', width = 1.05, fill ="#1f78b4") +
  scale_x_date(expand = c(0,0)) +
  scale_y_continuous(labels = fmt_dcmls(0), expand = c(0,0)) +
  theme(
    legend.position = "none",
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = cpal("light grey"), size = 0.21),
    panel.border = element_rect(fill = NA, color = cpal("border grey"), size = 0.3),
    plot.title = element_text(size = 21, color = "black", family = "Open Sans Light"),
    plot.subtitle = element_text(size = 14, color = cpal("text"), family = "Open Sans"),
    plot.caption = element_text(size = 12.5, color = cpal("text"), family = "Open Sans"),
    text = element_text(size = 15, color = cpal("text"), family = "Open Sans"),
    legend.text = element_text(size = 15, color = cpal("text"), family = "Open Sans"),
    axis.text.x = element_text(size = 14, color = cpal("text"), family = "Open Sans",
                               margin = unit(c(0.35, 0.25, 0.25, 0.25), "cm")),
    axis.text.y = element_text(size = 14, color = cpal("text"), family = "Open Sans",
                               margin = unit(c(0.35, 0.25, 0.25, 0.25), "cm")),
    axis.title.y = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.length = unit(-0.15, "cm"),
    axis.ticks = element_line(color = cpal("border grey"), size = 0.3),
    plot.margin = unit(c(0.3, 0.5, 0.3, 0.05), "cm")
  ) + 
  labs(title = ctitle_1,
       caption = paste("Data as of ", format(head(dat$date,1), "%b %d, %Y"), ". ", sep = "")) +
  transition_states(date) +
  shadow_mark()


anim1 <- animate(chart1, renderer = gifski_renderer(loop = T), fps = 50, duration = 7,
                 width = 300, height = 300, type = "cairo", end_pause = 85)  

  
ctitle_2 <- "Test positivity rate, %"

chart2 <- dat %>% 
  filter(date > "2020-04-30") %>% 
  ggplot(aes(x = date, y = percPositive)) +
  geom_line(size = 0.75, color ="#e31a1c", alpha = 0.65) +
  geom_point(color ="#e31a1c", size = 2.5) +
  scale_x_date(expand = c(0,0)) +
  scale_y_continuous(labels = fmt_dcmls(0), expand = c(0,0)) +
  theme(
    legend.position = "none",
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = cpal("light grey"), size = 0.21),
    panel.border = element_rect(fill = NA, color = cpal("border grey"), size = 0.3),
    plot.title = element_text(size = 21, color = "black", family = "Open Sans Light"),
    plot.subtitle = element_text(size = 14, color = cpal("text"), family = "Open Sans"),
    plot.caption = element_text(size = 12.5, color = cpal("text"), family = "Open Sans"),
    text = element_text(size = 15, color = cpal("text"), family = "Open Sans"),
    legend.text = element_text(size = 15, color = cpal("text"), family = "Open Sans"),
    axis.text.x = element_text(size = 14, color = cpal("text"), family = "Open Sans",
                               margin = unit(c(0.35, 0.25, 0.25, 0.25), "cm")),
    axis.text.y = element_text(size = 14, color = cpal("text"), family = "Open Sans",
                               margin = unit(c(0.35, 0.25, 0.25, 0.25), "cm")),
    axis.title.y = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.length = unit(-0.15, "cm"),
    axis.ticks = element_line(color = cpal("border grey"), size = 0.3),
    plot.margin = unit(c(0.3, 0.5, 0.3, 0.05), "cm")
  ) + 
  labs(title = ctitle_2,
       caption = paste(source_1[1], sep = "")) +
  transition_reveal(date) +
  shadow_mark()


anim2 <- animate(chart2, renderer = gifski_renderer(loop = T), fps = 50, duration = 7,
                 width = 300, height = 300, type = "cairo", end_pause = 85)  



a_mgif <- image_read(anim1)
b_mgif <- image_read(anim2)

new_gif <- image_append(c(a_mgif[1], b_mgif[1]))
for(i in 2:350){
  combined <- image_append(c(a_mgif[i], b_mgif[i]))
  new_gif <- c(new_gif, combined)
}

new_gif
###
```

California, Texas, Florida, and Georgia now appear well past the apex in the pace of new cases they report each day, though each continues to see daily increases in cases larger than those of the major European countries referenced above.

```{r cases, echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE}
#https://covidtracking.com/data/download/all-states-history.csv

raw_dat <- read_csv("D:\\Documents\\R directory\\covid19\\daily.csv")

dat <- raw_dat %>% 
  as_data_frame() %>%
  select(date, state, positive, negative, hospitalizedCumulative, death) %>% 
  mutate(date = as.Date(as.character(date), "%Y%m%d"),
         percPositive = (positive / (positive + negative)) * 100,
         totalTests = positive + negative)
  
dat_w <- dat %>% 
  select(date, state, positive) %>% 
  pivot_wider(names_from = "state", values_from = "positive") %>% 
  select(-GU, -DC, -PR, -VI, -MP, -AS) %>% 
  arrange(date)

rolling_mean <- rollify(mean, window = 7)
  
dat_avg <- dat_w[,-1] %>% 
  map(.f = function (x) {x - lag(x, 1)}) %>%
  map(.f = function (x) {rolling_mean(x)}) %>%
  as_data_frame() %>% 
  cbind(dat_w$date) %>% 
  rename(date = 'dat_w$date') %>% 
  select(date, everything()) %>% 
  as_data_frame()
  
dat_l <- dat_avg %>% 
  pivot_longer(names_to = "key", values_to = "value", -date)


dat_flag <- dat_l %>% 
  arrange(key, date) %>% 
  mutate(crossover = case_when((value >= 20 & lag(value,1) < 20) | 
                               (key != lag(key,1))~ 1,
                               TRUE ~ 0)) %>% 
  filter(key != "MI" & key != "ME")



dat_flag[dat_flag$date == "2020-03-19" & dat_flag$key == "MA", "crossover"] <- 1
dat_flag[dat_flag$date == "2020-03-11" & dat_flag$key == "NY", "crossover"] <- 1
dat_flag[dat_flag$date == "2020-07-08" & dat_flag$key == "HI", "crossover"] <- 0
dat_flag[dat_flag$date == "2020-04-23" & dat_flag$key == "ME", "crossover"] <- 0
dat_flag[dat_flag$date == "2020-04-25" & dat_flag$key == "ME", "crossover"] <- 0
dat_flag[dat_flag$date == "2020-07-19" & dat_flag$key == "ME", "crossover"] <- 0
dat_flag[dat_flag$date == "2020-06-25" & dat_flag$key == "MT", "crossover"] <- 0
dat_flag[dat_flag$date == "2020-07-07" & dat_flag$key == "NH", "crossover"] <- 0
dat_flag[dat_flag$date == "2020-07-12" & dat_flag$key == "NH", "crossover"] <- 0
dat_flag[dat_flag$date == "2020-03-26" & dat_flag$key == "OR", "crossover"] <- 0
dat_flag[dat_flag$date == "2020-05-13" & dat_flag$key == "WV", "crossover"] <- 0
dat_flag[dat_flag$date == "2020-05-20" & dat_flag$key == "WV", "crossover"] <- 0
dat_flag[dat_flag$date == "2020-06-14" & dat_flag$key == "WV", "crossover"] <- 0
dat_flag[dat_flag$date == "2020-06-19" & dat_flag$key == "WY", "crossover"] <- 0
dat_flag[dat_flag$date == "2020-06-22" & dat_flag$key == "WY", "crossover"] <- 0
dat_flag[dat_flag$date == "2020-07-25" & dat_flag$key == "ME", "crossover"] <- 0
dat_flag[dat_flag$date == "2020-08-27" & dat_flag$key == "NH", "crossover"] <- 0



##dat_flag %>% filter(crossover == 1) %>% view()


dat_count <- dat_flag %>%
  group_by(key, idx = cumsum(crossover == 1L)) %>% 
  mutate(counter = row_number(),
         counter = as.numeric(counter),
         phase = case_when((idx %% 2) == 0 ~ "off",
                           (idx %% 2) != 0 ~ "on")) %>%
  ungroup() %>% 
  mutate(days = case_when(phase == "off" ~ 0,
                          phase == "on" ~ counter)) %>% 
  select(-crossover, -idx, -counter, -phase) %>% 
  filter(days > 0) %>% 
  mutate(days = days - 1)

##dat_count %>% view()

state_reg <- read.csv("D:\\Documents\\R directory\\covid19\\state_region.csv") %>% 
  as_tibble()

dat_fin <- merge(dat_count, state_reg, by = "key")

dat_c <- dat_fin 
dat_lab <- dat_c %>% filter(key == "NY" | key == "NJ" | key == "MA" | key == "CA" | 
                            key == "AZ" | key == "FL" | key == "WA" |
                            key == "TX" | key == "GA" | 
                            key == "MI")



fmt_dcmls <- function(decimals = 0) {
  function(x) as.character(format(round(as.numeric(x), 0), nsmall=0, big.mark=","))
}

ctitle_1 <- "COVID-19 daily new cases"
unit_label <- "Daily change in cases, 7 day moving average"
source_1 <- "Source: The COVID Tracking Project, @benbakkum." 

line_cols <- c("NY" = "#1f78b4", "NJ" = "#1f78b4", "MA" = "#1f78b4", "CT" = "#CDCDCD", "PA" = "#CDCDCD",
               "CA" = "#33a02c", "AZ" = "#33a02c", "WA" = "#33a02c",
               "TX" = "#ff7f00", "GA" = "#ff7f00", "LA" = "#CDCDCD", "AL" = "#CDCDCD", "FL" = "#ff7f00", "TN" = "#CDCDCD",
               "MI" = "#6a3d9a",
               "AK" = "#CDCDCD", "CO" = "#CDCDCD", "HI" = "#CDCDCD", "ID" = "#CDCDCD", "MT" = "#CDCDCD", "NM" = "#CDCDCD",
               "NV" = "#CDCDCD", "OR" = "#CDCDCD", "UT" = "#CDCDCD", "WY" = "#CDCDCD", "IA" = "#CDCDCD", "IL" = "#CDCDCD",
               "IN" = "#CDCDCD", "KS" = "#CDCDCD", "MN" = "#CDCDCD", "MO" = "#CDCDCD", "ND" = "#CDCDCD", "NE" = "#CDCDCD",
               "OH" = "#CDCDCD", "OK" = "#CDCDCD", "SD" = "#CDCDCD", "WI" = "#CDCDCD", "AR" = "#CDCDCD", "KY" = "#CDCDCD",
               "MS" = "#CDCDCD", "NC" = "#CDCDCD", "SC" = "#CDCDCD", "VA" = "#CDCDCD", "WV" = "#CDCDCD", "DE" = "#CDCDCD",
               "MD" = "#CDCDCD", "ME" = "#CDCDCD", "NH" = "#CDCDCD", "RI" = "#CDCDCD", "VT" = "#CDCDCD")
               

chart1 <- ggplot(data = dat_c, aes(x = days, y = value, color = key, label = format(date, '%b %d'))) +
  geom_text(data = filter(dat_c, key == "WA"), aes(165, 33), color = "#808080", size = 13, alpha = .8, 
            family = "Open Sans") +
  geom_line(size = 1.1, alpha = 0.65) +
  scale_y_log10(labels = fmt_dcmls(0), breaks = c(20, 50, 100, 200, 500, 1000, 2000, 5000,
                                                  10000, 20000),
                limits = c(20, 20000),
                expand = c(0,0)) +
  scale_color_manual(values = line_cols) +
  geom_point(data = dat_lab, aes(x = days, y = value), size = 2.7) +
  geom_text_repel(data = dat_lab, aes(x = days, y = value, label = key), nudge_x = 3.5, nudge_y = 0.15,
                                show.legend = FALSE, size = 4, family = "Open Sans", fontface = "bold") +
  scale_x_continuous(limits = c(0, 190), expand = c(0,0)) +
  theme(
    legend.position = "none",
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = cpal("light grey"), size = 0.21),
    panel.border = element_rect(fill = NA, color = cpal("border grey"), size = 0.3),
    plot.title = element_text(size = 21, color = "black", family = "Open Sans Light"),
    plot.subtitle = element_text(size = 14, color = cpal("text"), family = "Open Sans"),
    plot.caption = element_text(size = 12.5, color = cpal("text"), family = "Open Sans"),
    text = element_text(size = 15, color = cpal("text"), family = "Open Sans"),
    legend.text = element_text(size = 15, color = cpal("text"), family = "Open Sans"),
    axis.text.x = element_text(size = 14, color = cpal("text"), family = "Open Sans",
                               margin = unit(c(0.35, 0.25, 0.25, 0.25), "cm")),
    axis.text.y = element_text(size = 14, color = cpal("text"), family = "Open Sans",
                               margin = unit(c(0.35, 0.25, 0.25, 0.25), "cm")),
    axis.title.y = element_blank(),
    axis.title.x = element_text(size = 14, color = cpal("text"), family = "Open Sans",
                               margin = unit(c(0.05, 0.05, 0.05, 0.05), "cm")),
    axis.ticks.length = unit(-0.15, "cm"),
    axis.ticks = element_line(color = cpal("border grey"), size = 0.3),
    plot.margin = unit(c(0.3, 0.5, 0.3, 0.05), "cm")
  ) + 
  labs(title = ctitle_1, subtitle = unit_label, x = "Number of days since averaging over 20 new cases a day",
       caption = paste("Data as of ", format(tail(dat_c$date,1), "%b %d, %Y"), ". ", source_1[1], sep = "")) +
  transition_reveal(date, keep_last = TRUE) +
  ease_aes('linear')
  
anim1 <- animate(chart1, renderer = gifski_renderer(loop = T), nframe = 475,
                 width = 615, height = 420, type = "cairo", end_pause = 200)  

anim1

```





