---
title: "Jobs -> The start of a stronger upswing"
date: 2021-04-02T21:13:14-05:00
draft: false
---
&nbsp;
&nbsp;

The labor market recovery picked up across the board in March, with over 900k nonfarm payrolls added to the economy. Un- and under-employment metrics fell while labor force participation rates (LFPRs) and employment to population ratios (EPOPs) rose. Such broad based improvement hadn't occurred since October, before the winter COVID surge, when the unemployment rate was a full percentage point higher. The colored shading in the table below highlights the changes in a given metric from the month before.


```{r setup, include=FALSE}

library(tidyverse)
library(blscrapeR)
library(janitor)
library(kableExtra)
library(formattable)
library(gganimate)
library(quantmod)
library(scales)
library(magick)
library(Cairo)

windowsFonts("Open Sans" = windowsFont("Open Sans"))

colors <- c('aqua' = "#45BCC9", 'green' = "#648C2E", 'deep blue' = "#00567D",
            'espresso' = "#310008", 'sand' = "#E1D1A7", 'orange' = "#F38B00",
            'red' = "#A51D31", 'dark grey' = "8E8279", 'border grey' = "#808080",
            'text' = "#222222", 'light grey' = "#CDCDCD")

cpal <- function(...) {
  cols <- c(...)
  if (is.null(cols))
    return(colors)
  colors[[cols]]
}

```


```{r echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE}

dat <- bls_api(c("LNS14000000", "LNS14000060", "LNS11300000", "LNS11300060", "LNS12300000", "LNS12300060", "LNS13327709",
                "LNS12032194", "LNS12000000", "LNS13008636", "LNS11000000", "CES0500000003", "CES0000000001"), 
              startyear = 2018, 
              endyear = 2021, 
              Sys.getenv("BLS_KEY"))

data.frame(name = c("UR", "Prime_age_UR", "LFPR", "Prime_age_LFPR", "EPOP", "Prime_age_EPOP", "U6", 
                    "PTER", "Employment", "Long_term_unemployed", "LF", "AHE", "PAYEMS"),
           code = c("LNS14000000", "LNS14000060", "LNS11300000", "LNS11300060", "LNS12300000", "LNS12300060", "LNS13327709",
                "LNS12032194", "LNS12000000", "LNS13008636", "LNS11000000", "CES0500000003", "CES0000000001")) -> key
```

```{r table, echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE, results='asis'}
dat %>% 
  mutate(date = paste0(periodName, " ", year, " 01"),
         date = as.Date(date, "%B %Y %d")) %>% 
  rename(code = seriesID) %>% 
  left_join(key, by = "code") %>% 
  select(value, date, name) %>% 
  pivot_wider(names_from = "name", values_from = "value") %>% 
  mutate(PTER_emp = round((PTER / Employment) * 100, 1),
         Long_term_UR = round((Long_term_unemployed / LF) * 100, 1),
         AHE_mm = round((AHE / lead(AHE, 1) - 1) * 100, 1),
         AHE_yy = round((AHE / lead(AHE, 12) - 1) * 100, 1),
         nfp_chg = PAYEMS - lead(PAYEMS, 1),
         ) %>% 
  select(date, nfp_chg, UR, Prime_age_UR, U6, LFPR, Prime_age_LFPR, EPOP, Prime_age_EPOP, 
         PTER_emp, Long_term_UR, AHE_mm, AHE_yy) %>% 
  head(11) %>% 
  arrange(date) %>% 
  mutate(date = format(date, "%b-%y")) %>% 
  t() %>% 
  as_tibble(rownames = "rowname") %>% 
  row_to_names(row_number = 1) -> df_tbl

df_tbl[,-1] %>% 
  mutate_if(is.character, as.numeric) %>% 
  bind_cols(df_tbl$date) %>% 
  rename(date = ...12) %>% 
  select(date, everything()) %>% 
  mutate(m1_chg = .[[3]] - .[[2]],
         m2_chg = .[[4]] - .[[3]],
         m3_chg = .[[5]] - .[[4]],
         m4_chg = .[[6]] - .[[5]],
         m5_chg = .[[7]] - .[[6]],
         m6_chg = .[[8]] - .[[7]],
         m7_chg = .[[9]] - .[[8]],
         m8_chg = .[[10]] - .[[9]],
         m9_chg = .[[11]] - .[[10]],
         m10_chg = .[[12]] - .[[11]]) -> df_chgs

df_chgs[,c(1,3,13,4,14,5,15,6,16,7,17,8,18,9,19,10,20,11,21,12,22)] %>% 
  rename(" " = date) %>% 
  as.data.frame() -> df_tbl

## row names
df_tbl[1,1] <- "NFP growth, k"
df_tbl[2,1] <- "UR, %"
df_tbl[3,1] <- "Prime-age UR, %"
df_tbl[4,1] <- "U6, %"
df_tbl[5,1] <- "LFPR, %"
df_tbl[6,1] <- "Prime-age LFPR, %"
df_tbl[7,1] <- "EPOP, %"
df_tbl[8,1] <- "Prime-age EPOP, %"
df_tbl[9,1] <- "PTER, % of payrolls"
df_tbl[10,1] <- "Long-term UR, %"
df_tbl[11,1] <- "AHE, % m/m"
df_tbl[12,1] <- "AHE, % y/y"

df_spk <- df_tbl[,c(1, seq(2,ncol(df_tbl), 2))]


## top row
fmt_dcmls <- function(x) {
   format(round(as.numeric(x), 0), nsmall=0, big.mark=",")
}

for (i in seq(2,(ncol(df_tbl)-1),2)) {
  df_tbl[1,i] <- fmt_dcmls(df_tbl[1,i])
}

hlight <- function(x) {
  y <- as.numeric(gsub(",", "", x))

  case_when(y > 500              ~ color_tile("#75e86d", "#75e86d")(x),
            y < 500 & y > 0      ~ color_tile("#cfffcc", "#cfffcc")(x),
            y < 0   & y > -500   ~ color_tile("#ffd6d6", "#ffd6d6")(x),
            y < -500             ~ color_tile("#f59393", "#f59393")(x))
}

for (i in seq(2,(ncol(df_tbl)-1),2)) {
  df_tbl[1,i] <- hlight(df_tbl[1,i])
}


## all rows
fmt_1dcml <- function(x) {
  y <- format(round(as.numeric(x), 1), nsmall=1)
  x <- as.character(y)
  z <- as.numeric(str_sub(x, 1, 4))
  z
}

for (j in 3:nrow(df_tbl)) {
  for (i in 2:ncol(df_tbl)) {
    df_tbl[j,i] <- fmt_1dcml(df_tbl[j,i])
  }
}

## decline is good
dec_is_good <- function(i, j) {
  y <- as.numeric(df_tbl[j,i+1])

  case_when(y <  -0.1              ~ color_tile("#75e86d", "#75e86d")(df_tbl[j,i]),
            (y >= -0.1) & (y < 0)      ~ color_tile("#cfffcc", "#cfffcc")(df_tbl[j,i]),
            y ==    0              ~ df_tbl[j,i],
            (y <=  0.1) & (y > 0)      ~ color_tile("#ffd6d6", "#ffd6d6")(df_tbl[j,i]),
            y >   0.1              ~ color_tile("#f59393", "#f59393")(df_tbl[j,i]))
}

for (j in c(2, 3, 4, 9, 10)) {
  for (i in seq(2,(ncol(df_tbl)-1),2)) {
    df_tbl[j,i] <- dec_is_good(i, j)
  }
}


## increase is good
inc_is_good <- function(i, j) {
  y <- as.numeric(df_tbl[j,i+1])

  # case_when(y >    0.1              ~ color_tile("#75e86d", "#75e86d")(df_tbl[j,i]),
  #           (y <=   0.1) & (y > 0)      ~ color_tile("#cfffcc", "#cfffcc")(df_tbl[j,i]),
  #           y ==     0              ~ df_tbl[j,i],
  #           (y >=  -0.1) & (y < 0)      ~ color_tile("#ffd6d6", "#ffd6d6")(df_tbl[j,i]),
  #           y <   -0.1              ~ color_tile("#f59393", "#f59393")(df_tbl[j,i]))
  
    ifelse(y >    0.1,              color_tile("#75e86d", "#75e86d")(df_tbl[j,i]),
    ifelse((y <=   0.1) & (y > 0),  color_tile("#cfffcc", "#cfffcc")(df_tbl[j,i]),
    ifelse(y ==     0,             df_tbl[j,i],
    ifelse((y >=  -0.1) & (y < 0),  color_tile("#ffd6d6", "#ffd6d6")(df_tbl[j,i]),
            color_tile("#f59393", "#f59393")(df_tbl[j,i])))))
}

for (j in c(5, 6, 7, 8, 11, 12)) {
  for (i in seq(2,(ncol(df_tbl)-1),2)) {
    df_tbl[j,i] <- inc_is_good(i, j)
  }
}


## https://github.com/haozhu233/kableExtra/issues/217

df_tbl %>% 
  kable(escape = F,
        align = c('l', rep('r', 19))) %>% 
  kable_paper("hover", full_width = F, html_font = "Open Sans") %>% 
  remove_column(3) %>% 
  remove_column(4) %>% 
  remove_column(5) %>% 
  remove_column(6) %>% 
  remove_column(7) %>% 
  remove_column(8) %>% 
  remove_column(9) %>% 
  remove_column(10) %>% 
  remove_column(11) %>% 
  remove_column(12) %>% 
  remove_column(2) %>% 
  remove_column(2) %>% 
  remove_column(2) %>% 
  add_footnote("Source: BLS, @benbakkum.", notation = "none") -> table

table

```
&nbsp;

Labor force participation has a long way still to go to return to pre-pandemic levels, even on a demographic adjusted basis, but March's uptick may suggest that the recent flatlining in important metrics may be starting to let up as warm weather and vaccinations support economic activity.

```{r echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE}
dat %>% 
  mutate(date = paste0(periodName, " ", year, " 01"),
         date = as.Date(date, "%B %Y %d")) %>% 
  rename(code = seriesID) %>% 
  left_join(key, by = "code") %>% 
  select(value, date, name) %>% 
  pivot_wider(names_from = "name", values_from = "value") %>% 
  mutate(PTER_emp = round((PTER / Employment) * 100, 1),
         Long_term_UR = round((Long_term_unemployed / LF) * 100, 1),
         AHE_mm = round((AHE / lead(AHE, 1) - 1) * 100, 1),
         AHE_yy = round((AHE / lead(AHE, 12) - 1) * 100, 1),
         nfp_chg = PAYEMS - lead(PAYEMS, 1),
         ) %>% 
  select(date, nfp_chg, UR, Prime_age_UR, U6, LFPR, Prime_age_LFPR, EPOP, Prime_age_EPOP, 
         PTER_emp, Long_term_UR, AHE_mm, AHE_yy) -> chrt_dat


chrt_dat %>% 
  pivot_longer(!date, names_to = "key", values_to = "values") %>% 
  filter(date >= as.Date("2018-12-31")) %>% 
  mutate(key = case_when(key == "nfp_chg"         ~ "NFP growth, k",
                         key == "UR"              ~ "UR, %",
                         key == "Prime_age_UR"    ~ "Prime-age UR, %",
                         key == "U6"              ~ "U6, %",
                         key == "LFPR"            ~ "LFPR, %",
                         key == "Prime_age_LFPR"  ~ "Prime-age LFPR, %",
                         key == "EPOP"            ~ "EPOP, %",
                         key == "Prime_age_EPOP"  ~ "Prime-age EPOP, %",
                         key == "PTER_emp"        ~ "PTER, % of payrolls",
                         key == "Long_term_UR"    ~ "Long-term UR, %",
                         key == "AHE_mm"          ~ "AHE, % m/m",
                         key == "AHE_yy"          ~ "AHE, % y/y")) -> chrt_dat_lng

chrt_dat_lng$key <- factor(chrt_dat_lng$key, 
                           levels = c("NFP growth, k",
                         "UR, %",
                         "Prime-age UR, %",
                         "U6, %",
                         "LFPR, %",
                         "Prime-age LFPR, %",
                         "EPOP, %",
                         "Prime-age EPOP, %",
                         "PTER, % of payrolls",
                         "Long-term UR, %",
                         "AHE, % m/m",
                         "AHE, % y/y"))      


source_1 <- "Source: BLS, @benbakkum."

chart1 <- chrt_dat_lng %>% 
  ggplot(aes(date, values)) +
  geom_step(color = cpal("deep blue")) +
  scale_x_date(breaks = c(as.Date("2019-01-01"), as.Date("2020-01-01"), as.Date("2021-01-01")),
               date_labels = "%Y") +
  facet_wrap(~key, scales = "free_y", ncol = 4) +
  theme(
    legend.position = "none",
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = cpal("light grey"), size = 0.21),
    panel.border = element_rect(fill = NA, color = cpal("border grey"), size = 0.3),
    plot.title = element_text(size = 21, color = "black", family = "Open Sans Light"),
    plot.subtitle = element_text(size = 14, color = cpal("text"), family = "Open Sans"),
    plot.caption = element_text(size = 12.5, color = cpal("text"), family = "Open Sans"),
    text = element_text(size = 15, color = cpal("text"), family = "Open Sans"),
    legend.text = element_text(size = 15, color = cpal("text"), family = "Open Sans"),
    axis.text.x = element_text(size = 14, color = cpal("text"), family = "Open Sans",
                               margin = unit(c(0.35, 0.25, 0.25, 0.25), "cm")),
    axis.text.y = element_text(size = 14, color = cpal("text"), family = "Open Sans",
                               margin = unit(c(0.35, 0.25, 0.25, 0.25), "cm")),
    axis.title.y = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.length = unit(-0.15, "cm"),
    axis.ticks = element_line(color = cpal("border grey"), size = 0.3),
    plot.margin = unit(c(0.3, 0.5, 0.3, 0.05), "cm"),
    strip.text.x = element_text(size = 12, color = cpal("text"), family = "Open Sans")
  ) +
  labs(caption = source_1) +
  transition_reveal(date)
  
  
anim1 <- animate(chart1, renderer = gifski_renderer(loop = T), nframe = 225,
                 width = 615, height = 500, type = "cairo", end_pause = 200)  

anim1


                         
                         
                         
                         
                  
```

The rebound in employment, which had earlier appear to have stalled out at a net amount of job losses near the depths of the Global Financial Crisis (GFC) in the US, now looks to have inflected upwards. This may be an indication that the pace of this recovery will not stay stuck at the anemic speed of the recovery that began in 2010. Certainly the current level of fiscal and monetary policy accommodation will help more than the deficient stimulus and outright austerity of the 2010s. 

```{r PAYEMS, echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE}


raw_dat <- getSymbols('PAYEMS',src='FRED', auto.assign=F) 

dat <- raw_dat %>% 
  as_tibble(rownames = NA) %>% 
  rownames_to_column() %>% 
  rename(date = rowname,
         jobs = PAYEMS) %>% 
  mutate(date = as.Date(date, "%Y-%m-%d"))

dat_ph <- dat %>% 
  mutate(recess = case_when(date >= "1948-09-01" & date <= "1950-07-01" ~ "1948", 
                            date >= "1953-07-01" & date <= "1955-06-01" ~ "1953",
                            date >= "1957-04-01" & date <= "1959-04-01" ~ "1957",
                            date >= "1960-04-01" & date <= "1961-12-01" ~ "1960",
                            date >= "1974-10-01" & date <= "1976-02-01" ~ "1974",
                            date >= "1980-03-01" & date <= "1981-01-01" ~ "1980",
                            date >= "1981-07-01" & date <= "1983-11-01" ~ "1981",
                            date >= "1990-06-01" & date <= "1993-02-01" ~ "1990",
                            date >= "2001-02-01" & date <= "2005-02-01" ~ "2001",
                            date >= "2008-01-01" & date <= "2014-05-01" ~ "2008",
                            date >= "2020-02-01" ~ "2020"))

dat_long <- dat_ph %>% 
  group_by(recess) %>% 
  mutate(perc = (jobs / first(jobs) - 1) * 100) %>% 
  na.omit() %>% 
  mutate(counter = row_number() - 1) %>%
  ungroup() %>% 
  mutate(idx = ifelse(recess == "1948", counter, #update date here
               ifelse(recess == "1953", counter + 4,
               ifelse(recess == "1957", counter + 8,
               ifelse(recess == "1960", counter + 12,
               ifelse(recess == "1974", counter + 16,
               ifelse(recess == "1980", counter + 20,
               ifelse(recess == "1981", counter + 24,
               ifelse(recess == "1990", counter + 28,
               ifelse(recess == "2001", counter + 32,
               ifelse(recess == "2008", counter + 36,
               ifelse(recess == "2020" & date == "2020-02-01", counter + 106,
               ifelse(recess == "2020" & date == "2020-03-01", counter + 114,
               ifelse(recess == "2020" & date == "2020-04-01", counter + 148,
               ifelse(recess == "2020" & date == "2020-05-01", counter + 156,
               ifelse(recess == "2020" & date == "2020-06-01", counter + 164,
               ifelse(recess == "2020" & date == "2020-07-01", counter + 172, 
               ifelse(recess == "2020" & date == "2020-08-01", counter + 180, 
               ifelse(recess == "2020" & date == "2020-09-01", counter + 188, 
               ifelse(recess == "2020" & date == "2020-10-01", counter + 196, 
               ifelse(recess == "2020" & date == "2020-11-01", counter + 204, 
               ifelse(recess == "2020" & date == "2020-12-01", counter + 212, 
               ifelse(recess == "2020" & date == "2021-01-01", counter + 220, 
               ifelse(recess == "2020" & date == "2021-02-01", counter + 228,
               ifelse(recess == "2020" & date == "2021-03-01", counter + 236, NA))))))))))))))))))))))))) %>% 
  mutate(recess = paste0(recess, " "))
        

start_date <- head(dat_long$date,1)
end_date <- tail(dat_long$date,1)

line_cols <- c("1948 " = "#b2df8a", 
               "1953 " = "#fb9a99", 
               "1957 " = "#fdbf6f", 
               "1960 " = "#ff7f00", 
               "1974 " = "#cab2d6", 
               "1980 " = "#6a3d9a", 
               "1981 " = "#b15928", 
               "1990 " = "#33a02c",
               "2001 " = "#a6cee3", 
               "2008 " = "#1f78b4", 
               "2020 " = "#e31a1c")

fmt_dcmls <- function(decimals = 0) {
  function(x) as.character(format(round(as.numeric(x), 0), nsmall=0, big.mark=","))
}

ctitle_1 <- "Job losses during US post-war recessions"
unit_label <- "% difference from cycle's peak total number of jobs"
source_1 <- "Source: BLS, FRED, @benbakkum."
x_lab <- "Number of months after peak employment"

chart_1 <- dat_long %>%
  ggplot(aes(x = counter, y = perc, color = recess)) +
  geom_line(size = 1, alpha = 0.85) +
  geom_point(size = 2.2) +
  geom_text(data = filter(dat_long, recess == "2008 " | recess == "2020 " | recess == "2001 "), 
            aes(x = counter, y = perc, label = recess),
            nudge_x = 3.2, 
            nudge_y = 0.15,
            show.legend = FALSE, 
            size = 4, 
            family = "Open Sans", 
            fontface = "bold") +
  scale_color_manual(values = line_cols) +
  scale_x_continuous(limits = c(0, 85), expand = c(0,0)) +
  scale_y_continuous(limits = c(-16, 1), expand = c(0,0), labels = fmt_dcmls(0)) +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.border = element_rect(fill = NA, color = cpal("border grey"), size = 0.3),
    plot.title = element_text(size = 21, color = "black", family = "Open Sans Light"),
    plot.subtitle = element_text(size = 14, color = cpal("text"), family = "Open Sans"),
    plot.caption = element_text(size = 12.5, color = cpal("text"), family = "Open Sans"),
    text = element_text(size = 15, color = cpal("text"), family = "Open Sans"),
    legend.text = element_text(size = 15, color = cpal("text"), family = "Open Sans"),
    axis.text.x = element_text(size = 14, color = cpal("text"), family = "Open Sans",
                               margin = unit(c(0.35, 0.25, 0.25, 0.25), "cm")),
    axis.text.y = element_text(size = 14, color = cpal("text"), family = "Open Sans",
                               margin = unit(c(0.25, 0.25, 0.25, 0.25), "cm")),
    axis.title.y = element_blank(),
    axis.title.x = element_text(size = 14, color = cpal("text"), family = "Open Sans"),
    axis.ticks.length = unit(-0.15, "cm"),
    axis.ticks = element_line(color = cpal("border grey"), size = 0.3),
    panel.grid.major = element_line(color = cpal("light grey"), size = 0.21),
    plot.margin = unit(c(0.3, 0.5, 0.3, 0.3), "cm"),
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.key = element_blank(),
    legend.key.size = unit(0.37, "cm"),
    legend.background = element_blank(),
    legend.margin = margin(t = 0, unit = "cm"),
    legend.spacing.x = unit(0.2, 'cm'),
    legend.spacing.y = unit(0.2, 'cm')) +
  guides(color=guide_legend(nrow=2,byrow=TRUE)) +
  labs(title = ctitle_1, 
       subtitle = unit_label,
       caption = paste("Data as of ", format(tail(dat$date,1), "%b %Y"), ". ", source_1[1], sep = ""),
       x = x_lab) +
  transition_reveal(idx, keep_last = TRUE) +
  ease_aes('linear') +
  view_follow(fixed_x = TRUE)

anim <- animate(chart_1, renderer = gifski_renderer(loop = T), nframe = 360,
                 width = 615, height = 505, type = "cairo", end_pause = 120)

anim
```

The leisure and hospitality industry remains the hardest hit, with the course of its employment levels resembling the path of the overall labor force participation rate. The industry had the strongest gains in March, however, and will likely continue to lead to the upside. Transportation & warehousing, professional services, and financial industries are now not far off from their peak amounts of employment in early 2020. 

```{r industries, echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE}
codes <- c("USCONS", "MANEMP", "USMINE", "USWTRADE", "USTRADE", "CES4300000001", "CES4422000001", "CES6054000001",
           "USINFO", "USFIRE", "CES6056000001", "CES6561000001", "CES6562000001", "USLAH", "USSERV")
key <- c("Construction", "Manufacturing", "Mining & Logging", "Wholesale Trade", "Retail Trade",
           "Transport & Warehouse", "Utilities", "Professional Svcs", "Information", "Financial",
           "Admin & Waste Svcs", "Educational Svcs", "Health Care",
           "Leisure & Hospitality", "Other Services")
cat <- c("Goods-producing", "Goods-producing", "Goods-producing", "Service-providing", "Service-providing",
         "Service-providing", "Service-providing", "Service-providing", "Service-providing", "Service-providing",
         "Service-providing", "Service-providing", "Service-providing", "Service-providing", "Service-providing")
cat_map <- data.frame(key, cat)

raw_dat <- getSymbols(codes,src='FRED') 

dat <- merge(USCONS, MANEMP, USMINE, USWTRADE, USTRADE, CES4300000001, CES4422000001, CES6054000001,
           USINFO, USFIRE, CES6056000001, CES6561000001, CES6562000001, USLAH, USSERV)  %>% 
  as_tibble(rownames = NA) %>% 
  rownames_to_column() 

colnames(dat) <- c("date", key)


chart1 <- dat %>% 
  filter(date > "2018-12-31") %>% 
  pivot_longer(!date) %>% 
  filter(!name %in% c("Other Services", "Utilities", "Wholesale Trade")) %>% 
  mutate(date = as.Date(date, "%Y-%m-%d"),
         name = fct_reorder(name, value, function(.) min(.) / max(.))) %>% 
  ggplot(aes(date, value)) +
  geom_step(color = cpal("deep blue")) +
  scale_x_date(breaks = c(as.Date("2019-01-01"), as.Date("2020-01-01"), as.Date("2021-01-01")),
               date_labels = "%Y") +
  facet_wrap(~ name, scales = "free_y") +
   theme(
    legend.position = "none",
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = cpal("light grey"), size = 0.21),
    panel.border = element_rect(fill = NA, color = cpal("border grey"), size = 0.3),
    plot.title = element_text(size = 21, color = "black", family = "Open Sans Light"),
    plot.subtitle = element_text(size = 14, color = cpal("text"), family = "Open Sans"),
    plot.caption = element_text(size = 12.5, color = cpal("text"), family = "Open Sans"),
    text = element_text(size = 15, color = cpal("text"), family = "Open Sans"),
    legend.text = element_text(size = 15, color = cpal("text"), family = "Open Sans"),
    axis.text.x = element_text(size = 14, color = cpal("text"), family = "Open Sans",
                               margin = unit(c(0.35, 0.25, 0.25, 0.25), "cm")),
    axis.text.y = element_text(size = 14, color = cpal("text"), family = "Open Sans",
                               margin = unit(c(0.35, 0.25, 0.25, 0.25), "cm")),
    axis.title.y = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.length = unit(-0.15, "cm"),
    axis.ticks = element_line(color = cpal("border grey"), size = 0.3),
    plot.margin = unit(c(0.3, 0.5, 0.3, 0.05), "cm"),
    strip.text.x = element_text(size = 11, color = cpal("text"), family = "Open Sans")
  ) +
  labs(caption = source_1) +
  transition_reveal(date)

anim1 <- animate(chart1, renderer = gifski_renderer(loop = T), nframe = 225,
                 width = 615, height = 500, type = "cairo", end_pause = 200)  

anim1

```
An aggregate measure of labor market conditions, the Blanchflower-Levin employment gap, continues to edge lower. It's now clear that this labor market shock now differs from GFC in the relatively smaller amount of underemployment being experienced. As the Fed continues to explicitly state that it cares about more than just the unemployment rate, the degree to which the participation gap shown below in green closes into next year will likely have significant implications for shifts in forward guidance and policy.


```{r blanchflower, echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE}
library(ggforce)
library(lubridate)

raw_dat <- getSymbols(c('UNRATE', 'CIVPART', 'LNS12032194', 'CLF16OV') ,src='FRED', auto.assign=T) 

dat <- merge(UNRATE, CIVPART, LNS12032194, CLF16OV) %>% 
  as_tibble(rownames = NA) %>% 
  rownames_to_column() %>% 
  rename(date = rowname,
         UR = UNRATE,
         LFPR = CIVPART,
         PTER = LNS12032194,
         LF = CLF16OV) %>% 
  mutate(date = as.Date(date, "%Y-%m-%d"),
         date = ceiling_date(date, "month") - 1)

## NAIRU
raw_nairu <- getSymbols('NROU' ,src='FRED', auto.assign=T) 

dat_nairu <- NROU %>% 
  as_tibble(rownames = NA) %>% 
  rownames_to_column() %>% 
  rename(date = rowname,
         NAIRU = NROU) %>% 
  mutate(date = as.Date(date, "%Y-%m-%d")) %>% 
  filter(date < Sys.Date())


mthly_nairu <- seq(dat_nairu$date[1] + 1, 
                   tail(dat_nairu$date,1) + 1, 
                   by = "1 month") - 1

nairu_m <- data.frame(date = ceiling_date(mthly_nairu, "month") - 1, 
                      nairu = spline(dat_nairu, 
                                      method = "fmm", 
                                      xout=mthly_nairu)$y) %>% 
  as_data_frame()


## potential LFPR
cbo_dat <- read_csv("D:\\Documents\\R directory\\cbo_dat.csv") %>% 
  select(date, potential_lfpr) %>% 
  mutate(date = as.yearqtr(date, "%Yq%q"),
         date = as.Date(date, 1))
  
monthly <- seq(cbo_dat$date[1] + 1, tail(cbo_dat$date,1) + 1, by = "1 month") - 1
cbo_m <- data.frame(date = monthly, 
                    cbo_lfpr = spline(cbo_dat, 
                                      method = "fmm", 
                                      xout=monthly)$y) %>% 
  as_data_frame() %>% 
  filter(date < Sys.Date())


## merge and gaps calculated
dat_tot <- dat %>% 
  left_join(nairu_m, by = "date") %>% 
  left_join(cbo_m, by = "date") %>% 
  filter(date > "1979-12-31",
         date <= tail(dat$date,1)) %>% 
  mutate(PTER_adj = PTER * (5.48/8.08),
         PTER_perc = PTER_adj / LF)

pter_avg <- dat_tot %>% 
  select(date, PTER_perc) %>% 
  filter(date > "1993-12-31",
         date <= "2007-12-31") %>% 
  summarize(avg = mean(PTER_perc)) %>% 
  unlist()

BL <- dat_tot %>% 
  mutate(PTER_gap = (PTER_perc - pter_avg) * 100,
         UR_gap = UR - nairu,
         LFPR_gap = cbo_lfpr - LFPR,
         total = PTER_gap + UR_gap + LFPR_gap) %>% 
  select(date, PTER_gap, UR_gap, LFPR_gap) %>%
  rename('Underemployment gap' = PTER_gap, 'Unemployment gap' = UR_gap, 'Pariticipation gap' = LFPR_gap) %>% 
  pivot_longer(cols = ends_with("gap"), names_to = "key", values_to = "value")

dat_line <- dat_tot %>% 
  mutate(PTER_gap = (PTER_perc - pter_avg) * 100,
         UR_gap = UR - nairu,
         LFPR_gap = cbo_lfpr - LFPR,
         total = PTER_gap + UR_gap + LFPR_gap)

ctitle_1 <- "Blanchflower-Levin employment gap"
unit_label <- "%"
source_1 <- "Source: BLS, FRED, @benbakkum."

start_date <- head(dat_line$date,1)
end_date <- tail(dat$date,1)

BL_chart <- BL %>% 
  mutate(placeholder = "X") %>% 
  ggplot() +
  geom_bar(aes(fill = key, y = value, x = date),
           position = "stack", 
           stat = "identity",
           width = 32) +
  facet_zoom(xy = date > "2018/8/1" & date < "2021/03/01", 
             zoom.size = .4) +
  geom_line(data = dat_line, aes(x = date, y = total), 
             size = 0.8, 
             alpha = .5, 
             linetype = "dashed") +
  scale_fill_manual(values = c("#33a02c", "#a6cee3", "#1f78b4")) +
  scale_y_continuous(expand = c(0,0)) +
  scale_x_date(date_labels = "%Y",
               expand = c(0,50)) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.key = element_blank(),
    legend.key.size = unit(0.37, "cm"),
    legend.background = element_blank(),
    legend.margin = margin(t = -0.5, unit = "cm"),
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = cpal("light grey"), size = 0.21),
    panel.border = element_rect(fill = NA, color = cpal("border grey"), size = 0.3),
    plot.title = element_text(size = 21, color = "black", family = "Open Sans Light"),
    plot.subtitle = element_text(size = 14, color = cpal("text"), family = "Open Sans"),
    plot.caption = element_text(size = 12.5, color = cpal("text"), family = "Open Sans"),
    text = element_text(size = 15, color = cpal("text"), family = "Open Sans"),
    legend.text = element_text(size = 12, color = cpal("text"), family = "Open Sans"),
    axis.text.x = element_text(size = 14, color = cpal("text"), family = "Open Sans",
                               margin = unit(c(0.35, 0.25, 0.25, 0.25), "cm")),
    axis.text.y = element_text(size = 14, color = cpal("text"), family = "Open Sans",
                               margin = unit(c(0.35, 0.25, 0.25, 0.25), "cm")),
    axis.title.y = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.length = unit(-0.15, "cm"),
    axis.ticks = element_line(color = cpal("border grey"), size = 0.3),
    plot.margin = unit(c(0.3, 0.5, 0.3, 0.05), "cm")
  ) + 
  labs(title = ctitle_1, subtitle = unit_label,
       caption = paste("Data as of ", format(tail(dat_line$date,1), "%b, %Y"), ". ", 
                       source_1[1], sep = "")) +
  transition_states(placeholder)

anim <- animate(BL_chart, renderer = gifski_renderer(loop = T), fps = 1, duration = 1,
                 width = 615, height = 420, type = "cairo")

anim
```

Note: the unemployment gap is the difference between the unemployment rate and the non-accelerating inflation rate of unemployment (NAIRU). The participation gap is the difference between the labor force participation rate (LFPR) and the CBO’s estimate of the potential LFPR. The underemployment gap is the difference between the number of employees working part-time for economic reasons as a percentage of the labor force, adjusted for the difference in the average number of hours worked by part-time and full-time employees, and the 1994-2007 average of this calculation.