---
title: "US Labor Market Update"
slug: "us-labor-market-update-august-2026"
date: 2026-08-07T12:00:00-05:00
draft: false
---
&nbsp;
&nbsp;



Nonfarm payrolls fell by 23k in July, the first monthly decline since February and well below the consensus estimate of a gain of about 80k. Government payrolls dropped by 53k, leisure and hospitality shed 40k, and retail lost 19k. Health care, the engine of job growth for the past two years, added just 22k, a fraction of its recent pace. The BLS also revised down its estimate for May by 66k (to a gain of 63k) and for June by 37k (to a gain of 20k), leaving the two months combined 103k lower than previously reported.

The chart below shows the evolution of estimates for monthly job gains/losses and later revisions (the "Last" estimate can represent either the final monthly revision or the annual benchmark revision, which tries to address drift in the sample of employers surveyed from reality due to business births and deaths). If anything, this chart illustrates the monthly payrolls figures' ability to serve as a random number generator, being both difficult to forecast and heavily subject to revision.
```{r setup, include=FALSE}

## --- Libraries ---------------------------------------------------------------

library(tidyverse)
library(blscrapeR)
library(janitor)
library(kableExtra)
library(formattable)
library(gganimate)
library(quantmod)
library(scales)
library(magick)
library(Cairo)
library(lubridate)
library(Rblpapi)
library(showtext)


## --- Fonts -------------------------------------------------------------------

font_add_google("Open Sans", "Open Sans")
font_add_google("Open Sans", "Open Sans Light", regular.wt = 300)
showtext_auto()


## --- Color palette -----------------------------------------------------------

colors <- c(
  "paleblue"    = "#a6cee3",
  "blue"        = "#1f78b4",
  "green"       = "#33a02c",
  "palegreen"   = "#b2df8a",
  "red"         = "#e31a1c",
  "palered"     = "#fb9a99",
  "orange"      = "#ff7f00",
  "paleorange"  = "#fdbf6f",
  "purple"      = "#6a3d9a",
  "palepurple"  = "#cab2d6",
  "border grey" = "#808080",
  "brown"       = "#b15928",
  "text"        = "#222222",
  "light grey"  = "#CDCDCD"
)

cpal <- function(...) {
  cols <- c(...)
  if (is.null(cols))
    return(colors)
  colors[[cols]]
}


## --- Date references ---------------------------------------------------------
## These update automatically when the file is re-rendered.

current_yr    <- year(Sys.Date())
current_month <- floor_date(Sys.Date(), "month")

## x-axis breaks: Jan 1 of each year from 7 years ago through current year
yr_breaks <- seq.Date(
  from = as.Date(paste0(current_yr - 7, "-01-01")),
  to   = as.Date(paste0(current_yr, "-01-01")),
  by   = "year"
)

## Chart lookback start (for filtering industry / small multiples data)
chart_start <- as.Date(paste0(current_yr - 7, "-12-31"))

## Pre-covid reference point (fixed historical anchor)
pre_covid_ref <- as.Date("2020-02-01")


## --- Shared chart label ------------------------------------------------------

source_1 <- "Source: BLS, @benbakkum."


## --- Shared number formatter -------------------------------------------------
## Returns a formatter function for use with scale_y_continuous labels.

fmt_dcmls <- function(decimals = 0) {
  function(x) {
    as.character(
      format(round(as.numeric(x), 0), nsmall = 0, big.mark = ",")
    )
  }
}


## --- Global ggplot theme -----------------------------------------------------

my_theme <- theme(
  legend.position   = "bottom",
  legend.title      = element_blank(),
  legend.key        = element_blank(),
  legend.key.size   = unit(0.37, "cm"),
  legend.background = element_blank(),
  legend.margin     = margin(t = -0.5, unit = "cm"),

  panel.background  = element_rect(fill = "white"),
  panel.grid.major  = element_line(
    color = cpal("light grey"),
    size  = 0.21
  ),
  panel.border = element_rect(
    fill  = NA,
    color = cpal("border grey"),
    size  = 0.3
  ),

  plot.title = element_text(
    size   = 16.2,
    color  = "black",
    family = "Open Sans Light"
  ),
  plot.subtitle = element_text(
    size   = 10.8,
    color  = cpal("text"),
    family = "Open Sans"
  ),
  plot.caption = element_text(
    size   = 9.6,
    color  = cpal("text"),
    family = "Open Sans"
  ),
  plot.margin = unit(c(0.3, 0.5, 0.3, 0.05), "cm"),

  text = element_text(
    size   = 11.6,
    color  = cpal("text"),
    family = "Open Sans"
  ),
  legend.text = element_text(
    size   = 10.4,
    color  = cpal("text"),
    family = "Open Sans",
    margin = margin(l = 2, r = 30, unit = "pt")
  ),

  axis.text.x = element_text(
    size   = 10.8,
    color  = cpal("text"),
    family = "Open Sans",
    margin = unit(c(0.35, 0.25, 0.25, 0.25), "cm")
  ),
  axis.text.y = element_text(
    size   = 10.8,
    color  = cpal("text"),
    family = "Open Sans",
    margin = unit(c(0.35, 0.25, 0.25, 0.25), "cm")
  ),
  axis.title.y      = element_blank(),
  axis.title.x      = element_blank(),
  axis.ticks.length  = unit(-0.15, "cm"),
  axis.ticks         = element_line(
    color = cpal("border grey"),
    size  = 0.3
  )
)
```


```{r monthly_jobs, echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE}

## --- Chart labels ---
ctitle_1  <- "Estimates of monthly change in nonfarm payrolls"
unit_label <- "Payrolls, thousands, seasonally adjusted"


## --- Pull NFP estimate data from Bloomberg ---
blpConnect()

bbg_start <- current_month %m-% months(10)
bbg_opts  <- c("RELEASE_DATE_OVERRIDE" = "1")

nfp_data <- bdh(
  "NFP TCH Index",
  c("BN_SURVEY_MEDIAN",
    "ACTUAL_RELEASE",
    "FIRST_REVISION",
    "PX_LAST"),
  start.date = bbg_start,
  overrides  = bbg_opts
)


## --- Wrangle NFP data ---
## Lag survey & actual so they align to the correct reporting month,
## keep only end-of-month rows, reshape to long format.
nfp_data %>%
  mutate(
    BN_SURVEY_MEDIAN = lead(BN_SURVEY_MEDIAN),
    ACTUAL_RELEASE   = lead(ACTUAL_RELEASE)
  ) %>%
  filter(date == ceiling_date(date, "month") - days(1)) %>%
  set_names(c("date", "Consensus", "1st", "2nd", "Last")) %>%
  pivot_longer(
    -date,
    values_to = "values",
    names_to  = "estimate"
  ) %>%
  filter(!is.na(values)) %>%

  ## Assign release dates: consensus & 1st come out same month,
  ## 2nd revision one month later, final ("Last") two months later.
  mutate(
    date_released = case_when(
      estimate == "1st"       ~ floor_date(date, "month"),
      estimate == "2nd"       ~ floor_date(date, "month") %m+% months(1),
      estimate == "Last"      ~ floor_date(date, "month") %m+% months(2),
      estimate == "Consensus" ~ floor_date(date, "month")
    ),
    released   = format(date_released, "%b '%y"),
    txt_labels = paste0(values)
  ) %>%

  ## Keep last ~9 months; drop revisions that haven't been released yet.
  filter(
    date > (ceiling_date(Sys.Date(), "month") - 1) %m-% months(9)
  ) %>%
  filter(
    !(estimate %in% c("2nd", "Last") & date == max(date)),
    !(estimate == "Last" &
        date == sort(unique(date), decreasing = TRUE)[2])
  ) %>%

  mutate(rank_index = as.double(dense_rank(as.numeric(date)))) %>%
  rename(Estimate = estimate) %>%

  ## Fill out the full grid of (date x Estimate) so every month
  ## has all four rows.
  complete(
    date, Estimate,
    fill = list(
      values     = NA_real_,
      txt_labels = NA_character_,
      released   = NA_character_
    )
  ) %>%

  ## Recompute rank_index and date_released for the newly added rows;
  ## set date_released to NA for unreleased estimates.
  mutate(rank_index = as.double(dense_rank(as.numeric(date)))) %>%
  mutate(
    date_released = case_when(
      is.na(values) & rank_index == max(rank_index)          ~ as.Date(NA),
      Estimate == "Last" & rank_index == max(rank_index) - 1 ~ as.Date(NA),
      Estimate %in% c("1st", "Consensus") ~ floor_date(date, "month"),
      Estimate == "2nd"  ~ floor_date(date, "month") %m+% months(1),
      Estimate == "Last" ~ floor_date(date, "month") %m+% months(2)
    ),
    released = format(date_released, "%b '%y")
  ) %>%

  ## Fill remaining NA date_released within each month, then set
  ## the penultimate "Last" to the max release date (appears last
  ## in the animation).
  group_by(date) %>%
  fill(date_released, .direction = "updown") %>%
  ungroup() %>%
  mutate(
    date_released = if_else(
      Estimate == "Last" & rank_index == max(rank_index) - 1,
      max(date_released, na.rm = TRUE),
      date_released
    ),
    released = format(date_released, "%b '%y")
  ) -> df


## --- Prep for plotting ---
df$Estimate <- factor(
  df$Estimate,
  levels = c("Consensus", "1st", "2nd", "Last")
)

## Reserve an empty band beneath the data for the big grey release-month
## watermark, sized from the data range so it can never sit on top of a
## point no matter how large that month's prints happen to be.
val_rng  <- range(df$values, na.rm = TRUE)
val_span <- diff(val_rng)

label_x  <- min(df$rank_index) + 1.2
label_y  <- val_rng[1] - 0.16 * val_span

## y limits must be computed with na.rm: df$values carries NAs for
## revisions that haven't been released yet.
y_limits <- c(label_y - 0.09 * val_span, val_rng[2] + 0.05 * val_span)


## --- Build animated chart ---
df %>%
  arrange(rank_index, Estimate) %>%
  mutate(
    val_label = if_else(is.na(values), "", paste0(values))
  ) %>%
  ggplot(aes(rank_index, values, color = Estimate, shape = Estimate)) +

  ## Points + text labels
  geom_point(size = 6, alpha = 0.6, na.rm = TRUE) +
  geom_text(
    data    = df %>% filter(Estimate == "1st"),
    aes(label_x, label_y, label = released),
    color   = "#808080",
    size    = 13,
    alpha   = 0.8,
    family  = "Open Sans"
  ) +
  geom_text(
    aes(rank_index, values, label = val_label),
    size        = 6,
    nudge_x     = 0.5,
    show.legend = FALSE
  ) +
  geom_hline(
    yintercept = 0,
    linetype   = "dashed",
    color      = cpal("border grey")
  ) +

  ## Scales
  scale_color_manual(values = c(
    "Consensus" = cpal("border grey"),
    "1st"       = cpal("blue"),
    "2nd"       = cpal("orange"),
    "Last"      = cpal("green")
  )) +
  scale_shape_manual(values = c(
    "Consensus" = 18,
    "1st"       = 16,
    "2nd"       = 17,
    "Last"      = 15
  )) +
  scale_x_continuous(
    breaks = seq(min(df$rank_index), max(df$rank_index)),
    labels = na.omit(unique(df$released[order(df$rank_index)]))
  ) +
  scale_y_continuous(
    limits = y_limits
  ) +

  ## Theme — only override legend.margin from my_theme
  my_theme +
  theme(legend.margin = margin(t = 0.1, unit = "cm")) +

  ## Labels + animation
  labs(
    title    = ctitle_1,
    subtitle = unit_label,
    caption  = source_1
  ) +
  transition_states(
    date_released,
    transition_length = 3,
    state_length      = 1,
    wrap              = FALSE
  ) +
  shadow_mark(exclude_layer = c(2, 3)) -> chart1


## --- Render animation ---
anim <- animate(
  chart1,
  renderer  = gifski_renderer(loop = TRUE),
  nframe    = 230,
  width     = 615,
  height    = 420,
  units     = "px",
  res       = 72,
  type      = "cairo",
  end_pause = 90
)

anim
```


The actual net change in payrolls in July, without adjusting for seasonal effects, was a decline of about 1.1 million. July is reliably a big down month as education payrolls roll off for the summer, but this one was about a quarter deeper than the trailing 10 year average for July of roughly 860k (excluding 2020).

The chart below shows the actual (non-seasonally adjusted) and seasonally adjusted figures for the last seven months, along with the rolling 10 year average for that month of the year and the average of the last four months.


```{r seasonals, echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE}

## --- Pull SA and NSA payroll data from BLS ---
## BLS_KEY is read from ~/.Renviron so the key stays out of the post
## source, which Hugo would otherwise publish verbatim.

dat <- bls_api(
  c("CEU0000000001", "CES0000000001"),
  startyear = current_yr - 16,
  endyear   = current_yr,
  Sys.getenv("BLS_KEY")
)

## Map series codes to friendly names
data.frame(
  name = c("PAYNSA", "PAYEMS"),
  code = c("CEU0000000001", "CES0000000001")
) -> key


## --- Compute monthly payroll changes (SA and NSA) ---
dat %>%
  mutate(
    date = paste0(periodName, " ", year, " 01"),
    date = as.Date(date, "%B %Y %d")
  ) %>%
  rename(code = seriesID) %>%
  left_join(key, by = "code") %>%
  select(value, date, name) %>%
  pivot_wider(names_from = "name", values_from = "value") %>%
  mutate(
    nfp_chg_sa  = PAYEMS - lead(PAYEMS, 1),
    nfp_chg_nsa = PAYNSA - lead(PAYNSA, 1)
  ) -> dat_chg


## --- Compute recent averages (last 4 months) ---
dat_chg[1:4, ] %>%
  select(nfp_chg_nsa, nfp_chg_sa) %>%
  summarize(
    avg_sa  = mean(nfp_chg_sa),
    avg_nsa = mean(nfp_chg_nsa)
  ) -> avgs_curr_yr


## --- Compute trailing 10-year monthly averages ---
## 2020 is excluded: the pandemic collapse and rebound (April -19.7M NSA,
## June +5.1M) are one-off outliers that swamp a 10-observation mean --
## with them in, April's "average" reads -1.1M against actual Aprils of
## roughly +1M. The chart labels the exclusion. Revisit in 2030 when
## 2020 rolls out of the window anyway.
dat_chg %>%
  filter(date > (max(date) %m-% months(120))) %>%
  filter(year(date) != 2020) %>%
  select(date, nfp_chg_nsa, nfp_chg_sa) %>%
  mutate(month = month(date)) %>%
  group_by(month) %>%
  summarize(
    nfp_chg_sa  = mean(nfp_chg_sa),
    nfp_chg_nsa = mean(nfp_chg_nsa)
  ) %>%
  pivot_longer(
    -month,
    names_to  = "key",
    values_to = "avg_10_yr"
  ) -> avgs_10_yr


## --- Merge actuals with both averages, filter to recent window ---
dat_chg %>%
  select(date, nfp_chg_sa, nfp_chg_nsa) %>%
  pivot_longer(-date, names_to = "key", values_to = "values") %>%
  mutate(month = month(date)) %>%
  left_join(avgs_10_yr, by = c("month", "key")) %>%
  mutate(
    avg_curr_yr = case_when(
      key == "nfp_chg_sa" &
        date > (ceiling_date(Sys.Date(), "month") - 1) %m-% months(5)
        ~ avgs_curr_yr$avg_sa,
      key == "nfp_chg_nsa" &
        date > (ceiling_date(Sys.Date(), "month") - 1) %m-% months(5)
        ~ avgs_curr_yr$avg_nsa
    )
  ) %>%
  filter(
    date > (ceiling_date(Sys.Date(), "month") - 1) %m-% months(8)
  ) -> dat_chg_avg


## --- Reshape to long format for plotting ---
## Combines actuals, 10y avgs, and current avgs into a single
## key/values column with names like avg_10_sa, avg_curr_nsa, etc.
dat_chg_avg %>%
  pivot_longer(
    cols      = c(values, avg_10_yr, avg_curr_yr),
    names_to  = "metric",
    values_to = "values"
  ) %>%
  mutate(
    key = case_when(
      metric == "values"      ~ key,
      metric == "avg_10_yr"   ~ str_replace(key, "nfp_chg", "avg_10"),
      metric == "avg_curr_yr" ~ str_replace(key, "nfp_chg", "avg_curr")
    )
  ) %>%
  select(date, key, values) %>%
  mutate(placeholder = "x") -> dat_final


## --- Chart labels ---
ctitle_1   <- "Seasonally and non-seasonally adjusted payroll gains"
unit_label <- "Change in jobs, thousands"


## --- Build chart ---
dat_final %>%
  filter(key %in% c("nfp_chg_sa", "nfp_chg_nsa")) %>%
  ggplot(aes(date, values, fill = key)) +

  ## Bars: SA vs NSA monthly changes
  geom_col(position = "dodge", alpha = 0.8) +

  ## Points: trailing 10-year average for each month
  geom_point(
    data     = dat_final %>% filter(key %in% c("avg_10_sa", "avg_10_nsa")),
    aes(date, values, shape = key, color = key),
    position = position_dodge(width = 25),
    size     = 3,
    alpha    = 0.6
  ) +

  ## Points: last four month average
  geom_point(
    data     = dat_final %>% filter(key %in% c("avg_curr_sa", "avg_curr_nsa")),
    aes(date, values, shape = key, color = key),
    position = position_dodge(width = 25),
    size     = 4,
    alpha    = 0.6
  ) +

  geom_hline(
    yintercept = 0,
    linetype   = "dashed",
    color      = cpal("border grey"),
    size       = 1
  ) +

  ## Scales
  scale_x_date(
    date_breaks = "1 month",
    date_labels = "%b '%y"
  ) +
  scale_y_continuous(labels = fmt_dcmls(0)) +

  scale_fill_manual(
    values = c(
      "nfp_chg_nsa" = cpal("blue"),
      "nfp_chg_sa"  = cpal("green")
    ),
    labels = c(
      "nfp_chg_nsa" = "Non-seasonally adjusted",
      "nfp_chg_sa"  = "Seasonally adjusted"
    )
  ) +
  scale_shape_manual(
    values = c(
      "avg_10_nsa"  = 15,
      "avg_10_sa"   = 15,
      "avg_curr_sa" = 18,
      "avg_curr_nsa" = 18
    ),
    labels = c(
      "Trailing 10y avg, excl. 2020",
      "Last four month avg"
    ),
    breaks = c("avg_10_nsa", "avg_curr_sa")
  ) +
  scale_color_manual(
    values = c(
      "avg_10_nsa"  = cpal("purple"),
      "avg_10_sa"   = cpal("purple"),
      "avg_curr_sa" = cpal("orange"),
      "avg_curr_nsa" = cpal("orange")
    ),
    labels = c(
      "Trailing 10y avg, excl. 2020",
      "Last four month avg"
    ),
    breaks = c("avg_10_nsa", "avg_curr_nsa")
  ) +

  ## Legend overrides to clean up combined fill/shape/color legend
  guides(
    fill = guide_legend(
      override.aes = list(
        fill  = c(cpal("blue"), cpal("green")),
        shape = c(NA, NA),
        color = c(NA, NA)
      )
    ),
    ncol = 1
  ) +

  ## Theme
  my_theme +
  theme(
    legend.direction = "vertical",
    legend.spacing.x = unit(2, "cm"),
    legend.text      = element_text(margin = margin(t = 3, b = 3)),
    axis.text.x      = element_text(
      size   = 10.8,
      margin = unit(c(0.35, 0.25, 0.25, 0.25), "cm"),
      angle  = 50,
      vjust  = 0.85,
      hjust  = 0.75
    )
  ) +

  labs(
    title    = ctitle_1,
    subtitle = unit_label,
    caption  = source_1
  ) +
  transition_manual(placeholder) -> p

anim1 <- animate(
  p,
  renderer = gifski_renderer(loop = TRUE),
  nframe   = 1,
  width    = 615,
  height   = 475,
  units    = "px",
  res      = 72,
  type     = "cairo"
)

anim1

```


A quick guide to the table and charts below:

- **NFP growth**: the monthly change in total nonfarm payrolls, in thousands.
- **UR**: the unemployment rate, i.e. the share of people in the labor force who don't have a job and are actively looking for one as a percentage of the labor force.
- **Prime-age UR**: the unemployment rate for 25-54 year olds, filtering out students and retirees. This to a large extent controls for changes in the UR due to demographic shifts.
- **U6**: the broader unemployment measure that also includes people who are marginally attached to the labor market, having looked for a job in the last 12 months but not in the last four weeks, and people stuck in part-time jobs who want full-time work.
- **LFPR**: labor force participation rate, the share of the civilian population that is either working or looking for work. The higher the labor force participation, the healthier the labor market.
- **EPOP**: the employment-to-population ratio. What share of the population has a job? This metric accounts for both employment and labor force participation in measuring job market conditions. That combo, along with stripping out some of the effects of demographics, makes prime-age EPOP arguably the most well rounded metric.
- **PTER**: part-time for economic reasons, people working part-time because they can't find full-time work, as a share of total payrolls.
- **Long-term UR**: the share of unemployed workers who've been out of a job for 27 weeks or more.
- **AHE**: average hourly earnings. How much workers are getting paid per hour, shown both month-over-month and year-over-year.

July's drop in the unemployment rate to 4.1% flatters the household survey: the labor force shrank by 264k and participation fell to 61.4%, its lowest in over five years, so fewer people counted as unemployed largely because fewer people counted at all. Wage growth was nearly flat on the month, with the year-over-year pace of average hourly earnings slipping to 3.2%, the slowest since May 2021. Prime-age EPOP ticking up two-tenths is the one genuinely firm-looking number in the set.


<link href="https://fonts.googleapis.com/css2?family=Open+Sans:wght@300;400;600;700&display=swap" rel="stylesheet">

```{r table, echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE, results='asis'}

## --- Pull household survey + establishment data from BLS ---

dat <- bls_api(
  c("LNS14000000", "LNS14000060",
    "LNS11300000", "LNS11300060",
    "LNS12300000", "LNS12300060",
    "LNS13327709", "LNS12032194",
    "LNS12000000", "LNS13008636",
    "LNS11000000", "CES0500000003",
    "CES0000000001"),
  startyear = current_yr - 8,
  endyear   = current_yr,
  Sys.getenv("BLS_KEY")
)

## Map series codes to friendly column names
data.frame(
  name = c("UR", "Prime_age_UR",
           "LFPR", "Prime_age_LFPR",
           "EPOP", "Prime_age_EPOP",
           "U6", "PTER",
           "Employment", "Long_term_unemployed",
           "LF", "AHE", "PAYEMS"),
  code = c("LNS14000000", "LNS14000060",
           "LNS11300000", "LNS11300060",
           "LNS12300000", "LNS12300060",
           "LNS13327709", "LNS12032194",
           "LNS12000000", "LNS13008636",
           "LNS11000000", "CES0500000003",
           "CES0000000001")
) -> key


## --- Build wide data with derived metrics ---
dat %>%
  mutate(
    date = paste0(periodName, " ", year, " 01"),
    date = as.Date(date, "%B %Y %d")
  ) %>%
  rename(code = seriesID) %>%
  left_join(key, by = "code") %>%
  select(value, date, name) %>%
  pivot_wider(names_from = "name", values_from = "value") %>%
  fill(everything(), .direction = "downup") %>%
  mutate(
    PTER_emp     = round((PTER / Employment) * 100, 1),
    Long_term_UR = round((Long_term_unemployed / LF) * 100, 1),
    AHE_mm       = round((AHE / lead(AHE, 1) - 1) * 100, 1),
    AHE_yy       = round((AHE / lead(AHE, 12) - 1) * 100, 1),
    nfp_chg      = PAYEMS - lead(PAYEMS, 1)
  ) %>%
  select(
    date, nfp_chg, UR, Prime_age_UR, U6,
    LFPR, Prime_age_LFPR, EPOP, Prime_age_EPOP,
    PTER_emp, Long_term_UR, AHE_mm, AHE_yy
  ) %>%
  head(11) %>%
  arrange(date) %>%
  mutate(date = format(date, "%b-%y")) %>%
  t() %>%
  as_tibble(rownames = "rowname") %>%
  row_to_names(row_number = 1) -> df_tbl


## --- Compute month-over-month changes for heat-map coloring ---
df_tbl[, -1] %>%
  mutate_if(is.character, as.numeric) %>%
  bind_cols(df_tbl$date) %>%
  rename(date = ...12) %>%
  select(date, everything()) %>%
  mutate(
    m1_chg  = .[[3]]  - .[[2]],
    m2_chg  = .[[4]]  - .[[3]],
    m3_chg  = .[[5]]  - .[[4]],
    m4_chg  = .[[6]]  - .[[5]],
    m5_chg  = .[[7]]  - .[[6]],
    m6_chg  = .[[8]]  - .[[7]],
    m7_chg  = .[[9]]  - .[[8]],
    m8_chg  = .[[10]] - .[[9]],
    m9_chg  = .[[11]] - .[[10]],
    m10_chg = .[[12]] - .[[11]]
  ) -> df_chgs

## Interleave value and change columns
df_chgs[, c(1, 3, 13, 4, 14, 5, 15, 6, 16, 7, 17,
            8, 18, 9, 19, 10, 20, 11, 21, 12, 22)] %>%
  rename(" " = date) %>%
  as.data.frame() -> df_tbl


## --- Row labels ---
df_tbl[1, 1]  <- "NFP growth, k"
df_tbl[2, 1]  <- "UR, %"
df_tbl[3, 1]  <- "Prime-age UR, %"
df_tbl[4, 1]  <- "U6, %"
df_tbl[5, 1]  <- "LFPR, %"
df_tbl[6, 1]  <- "Prime-age LFPR, %"
df_tbl[7, 1]  <- "EPOP, %"
df_tbl[8, 1]  <- "Prime-age EPOP, %"
df_tbl[9, 1]  <- "PTER, % of payrolls"
df_tbl[10, 1] <- "Long-term UR, %"
df_tbl[11, 1] <- "AHE, % m/m"
df_tbl[12, 1] <- "AHE, % y/y"

df_spk <- df_tbl[, c(1, seq(2, ncol(df_tbl), 2))]


## --- Format and color-code NFP row (top row) ---

fmt_dcmls_tbl <- function(x) {
  format(round(as.numeric(x), 0), nsmall = 0, big.mark = ",")
}

for (i in seq(2, (ncol(df_tbl) - 1), 2)) {
  df_tbl[1, i] <- fmt_dcmls_tbl(df_tbl[1, i])
}

## Apply heat-map coloring to NFP values
hlight <- function(x) {
  y <- as.numeric(gsub(",", "", x))

  case_when(
    y > 500            ~ color_tile("#75e86d", "#75e86d")(x),
    y < 500 & y > 0    ~ color_tile("#cfffcc", "#cfffcc")(x),
    y < 0   & y > -500 ~ color_tile("#ffd6d6", "#ffd6d6")(x),
    y < -500           ~ color_tile("#f59393", "#f59393")(x)
  )
}

for (i in seq(2, (ncol(df_tbl) - 1), 2)) {
  df_tbl[1, i] <- hlight(df_tbl[1, i])
}


## --- Format remaining rows to one decimal place ---

fmt_1dcml <- function(x) {
  y <- format(round(as.numeric(x), 1), nsmall = 1)
  x <- as.character(y)
  z <- as.numeric(str_sub(x, 1, 4))
  z
}

for (j in 3:nrow(df_tbl)) {
  for (i in 2:ncol(df_tbl)) {
    df_tbl[j, i] <- fmt_1dcml(df_tbl[j, i])
  }
}


## --- Color-code metrics where a decline is good ---
## (UR, Prime-age UR, U6, PTER, Long-term UR)

dec_is_good <- function(i, j) {
  y <- as.numeric(df_tbl[j, i + 1])

  case_when(
    y < -0.1                 ~ color_tile("#75e86d", "#75e86d")(df_tbl[j, i]),
    (y >= -0.1) & (y < 0)   ~ color_tile("#cfffcc", "#cfffcc")(df_tbl[j, i]),
    y == 0                   ~ df_tbl[j, i],
    (y <= 0.1) & (y > 0)    ~ color_tile("#ffd6d6", "#ffd6d6")(df_tbl[j, i]),
    y > 0.1                  ~ color_tile("#f59393", "#f59393")(df_tbl[j, i])
  )
}

for (j in c(2, 3, 4, 9, 10)) {
  for (i in seq(2, (ncol(df_tbl) - 1), 2)) {
    df_tbl[j, i] <- dec_is_good(i, j)
  }
}


## --- Color-code metrics where an increase is good ---
## (LFPR, Prime-age LFPR, EPOP, Prime-age EPOP, AHE m/m, AHE y/y)

inc_is_good <- function(i, j) {
  y <- as.numeric(df_tbl[j, i + 1])

  ifelse(y > 0.1,                color_tile("#75e86d", "#75e86d")(df_tbl[j, i]),
  ifelse((y <= 0.1) & (y > 0),  color_tile("#cfffcc", "#cfffcc")(df_tbl[j, i]),
  ifelse(y == 0,                 df_tbl[j, i],
  ifelse((y >= -0.1) & (y < 0), color_tile("#ffd6d6", "#ffd6d6")(df_tbl[j, i]),
                                 color_tile("#f59393", "#f59393")(df_tbl[j, i])))))
}

for (j in c(5, 6, 7, 8, 11, 12)) {
  for (i in seq(2, (ncol(df_tbl) - 1), 2)) {
    df_tbl[j, i] <- inc_is_good(i, j)
  }
}


## --- Render HTML table ---
## ref: https://github.com/haozhu233/kableExtra/issues/217

df_tbl %>%
  kable(
    format = "html",
    escape = FALSE,
    align  = c("l", rep("r", 19))
  ) %>%
  kable_paper("hover", full_width = FALSE, html_font = "Open Sans") %>%
  remove_column(3) %>%
  remove_column(4) %>%
  remove_column(5) %>%
  remove_column(6) %>%
  remove_column(7) %>%
  remove_column(8) %>%
  remove_column(9) %>%
  remove_column(10) %>%
  remove_column(11) %>%
  remove_column(12) %>%
  remove_column(2) %>%
  remove_column(2) %>%
  remove_column(2) %>%
  add_footnote("Source: BLS, @benbakkum.", notation = "none") -> table

table

```
&nbsp;



```{r metrics_facet, echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE}

## --- Compute derived metrics for small-multiples chart ---
dat %>%
  mutate(
    date = paste0(periodName, " ", year, " 01"),
    date = as.Date(date, "%B %Y %d")
  ) %>%
  rename(code = seriesID) %>%
  left_join(key, by = "code") %>%
  select(value, date, name) %>%
  pivot_wider(names_from = "name", values_from = "value") %>%
  mutate(
    PTER_emp     = round((PTER / Employment) * 100, 1),
    Long_term_UR = round((Long_term_unemployed / LF) * 100, 1),
    AHE_mm       = round((AHE / lead(AHE, 1) - 1) * 100, 1),
    AHE_yy       = round((AHE / lead(AHE, 12) - 1) * 100, 1),
    nfp_chg      = PAYEMS - lead(PAYEMS, 1)
  ) %>%
  select(
    date, nfp_chg, UR, Prime_age_UR, U6,
    LFPR, Prime_age_LFPR, EPOP, Prime_age_EPOP,
    PTER_emp, Long_term_UR, AHE_mm, AHE_yy
  ) -> chrt_dat


## --- Reshape to long format with display-friendly labels ---
chrt_dat %>%
  pivot_longer(!date, names_to = "key", values_to = "values") %>%
  filter(date >= chart_start) %>%
  mutate(
    key = case_when(
      key == "nfp_chg"        ~ "NFP growth, k",
      key == "UR"             ~ "UR, %",
      key == "Prime_age_UR"   ~ "Prime-age UR, %",
      key == "U6"             ~ "U6, %",
      key == "LFPR"           ~ "LFPR, %",
      key == "Prime_age_LFPR" ~ "Prime-age LFPR, %",
      key == "EPOP"           ~ "EPOP, %",
      key == "Prime_age_EPOP" ~ "Prime-age EPOP, %",
      key == "PTER_emp"       ~ "PTER, % of payrolls",
      key == "Long_term_UR"   ~ "Long-term UR, %",
      key == "AHE_mm"         ~ "AHE, % m/m",
      key == "AHE_yy"         ~ "AHE, % y/y"
    )
  ) -> chrt_dat_lng

## Set facet display order
chrt_dat_lng$key <- factor(
  chrt_dat_lng$key,
  levels = c(
    "NFP growth, k",
    "UR, %",
    "Prime-age UR, %",
    "U6, %",
    "LFPR, %",
    "Prime-age LFPR, %",
    "EPOP, %",
    "Prime-age EPOP, %",
    "PTER, % of payrolls",
    "Long-term UR, %",
    "AHE, % m/m",
    "AHE, % y/y"
  )
)


## --- Build animated small-multiples chart ---
chart1 <- chrt_dat_lng %>%
  filter(date > max(chrt_dat_lng$date) %m-% months(60)) %>%
  ggplot(aes(date, values)) +
  geom_step(color = cpal("blue")) +
  scale_x_date(breaks = yr_breaks, date_labels = "'%y") +
  facet_wrap(~key, scales = "free_y", ncol = 4) +
  my_theme +
  theme(
    legend.position = "none",
    strip.text.x = element_text(
      size   = 9,
      color  = cpal("text"),
      family = "Open Sans"
    ),
    axis.text.x = element_text(
      size   = 9,
      color  = cpal("text"),
      family = "Open Sans"
    )
  ) +
  labs(caption = source_1) +
  transition_reveal(date)

anim1 <- animate(
  chart1,
  renderer  = gifski_renderer(loop = TRUE),
  nframe    = 225,
  width     = 615,
  height    = 500,
  units     = "px",
  res       = 72,
  type      = "cairo",
  end_pause = 200
)

anim1
```


Looking at longer-term trends across industries, the post-pandemic labor market has been defined by a lopsided recovery. Health care has been the dominant driver of job growth among the industries below, while goods-producing sectors like manufacturing and mining have been flat to declining. Leisure and hospitality, which bore the brunt of pandemic layoffs, has largely recovered but job growth there has stalled. Information sector employment, which the BLS defines as publishing (including software), film and sound recording, broadcasting, telecom, and data processing, has trended lower since the over-hiring of the pandemic recovery. 


```{r industries, echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE}

## --- Pull industry-level payroll data from FRED ---

codes <- c(
  "USCONS",        "MANEMP",        "USMINE",
  "USWTRADE",      "USTRADE",       "CES4300000001",
  "CES4422000001", "CES6054000001", "USINFO",
  "USFIRE",        "CES6056000001", "CES6561000001",
  "CES6562000001", "USLAH",         "USSERV"
)

key <- c(
  "Construction",         "Manufacturing",
  "Mining & Logging",     "Wholesale Trade",
  ## "Transport & Whse" is abbreviated deliberately: the full
  ## "Transport & Warehouse" overruns the facet strip at this panel width.
  "Retail Trade",         "Transport & Whse",
  "Utilities",            "Professional Svcs",
  "Information",          "Financial",
  "Admin & Waste Svcs",   "Educational Svcs",
  "Health Care",          "Leisure & Hospitality",
  "Other Services"
)

cat <- c(
  "Goods-producing",   "Goods-producing",   "Goods-producing",
  "Service-providing", "Service-providing", "Service-providing",
  "Service-providing", "Service-providing", "Service-providing",
  "Service-providing", "Service-providing", "Service-providing",
  "Service-providing", "Service-providing", "Service-providing"
)

cat_map <- data.frame(key, cat)

raw_dat <- getSymbols(codes, src = "FRED")

dat <- merge(
  USCONS, MANEMP, USMINE, USWTRADE, USTRADE,
  CES4300000001, CES4422000001, CES6054000001,
  USINFO, USFIRE, CES6056000001, CES6561000001,
  CES6562000001, USLAH, USSERV
) %>%
  as_tibble(rownames = NA) %>%
  rownames_to_column()

colnames(dat) <- c("date", key)


## --- Build animated industry facet chart ---
## Filter to last 5 years; exclude low-signal industries.
chart1 <- dat %>%
  filter(date > max(chrt_dat_lng$date) %m-% months(60)) %>%
  pivot_longer(!date) %>%
  filter(
    !name %in% c("Other Services", "Utilities", "Wholesale Trade")
  ) %>%
  mutate(
    date = as.Date(date, "%Y-%m-%d"),
    name = fct_reorder(name, value, function(.) min(.) / max(.))
  ) %>%
  ggplot(aes(date, value)) +
  geom_step(color = cpal("blue")) +
  scale_x_date(breaks = yr_breaks, date_labels = "'%y") +
  facet_wrap(~name, scales = "free_y") +
  my_theme +
  theme(
    legend.position = "none",
    strip.text.x = element_text(
      size   = 8,
      color  = cpal("text"),
      family = "Open Sans"
    ),
    axis.text.x = element_text(
      size   = 8,
      color  = cpal("text"),
      family = "Open Sans"
    ),
    axis.text.y = element_text(
      size   = 7.5,
      color  = cpal("text"),
      family = "Open Sans"
    )
  ) +
  labs(caption = source_1) +
  transition_reveal(date)

anim1 <- animate(
  chart1,
  renderer  = gifski_renderer(loop = TRUE),
  nframe    = 225,
  width     = 615,
  height    = 500,
  units     = "px",
  res       = 72,
  type      = "cairo",
  end_pause = 200
)

anim1

```


The flattish job growth of 2025 and 2026 (so far) has widened the gap between current overall employment and levels in the unlikely case the pre-pandemic trend of job growth had continued.


```{r trajectory, echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE}

## --- Pull SA/NSA payrolls (wider range for trajectory projection) ---

raw_dat <- bls_api(
  c("CEU0000000001", "CES0000000001"),
  startyear = max(current_yr - 19, 2005),
  endyear   = current_yr,
  Sys.getenv("BLS_KEY")
)

data.frame(
  name = c("PAYNSA", "PAYEMS"),
  code = c("CEU0000000001", "CES0000000001")
) -> key


## --- Build actual payrolls with anchor points for projections ---
raw_dat %>%
  mutate(
    date = paste0(periodName, " ", year, " 01"),
    date = as.Date(date, "%B %Y %d")
  ) %>%
  rename(code = seriesID) %>%
  left_join(key, by = "code") %>%
  select(value, date, name) %>%
  pivot_wider(names_from = "name", values_from = "value") %>%
  select(-PAYNSA) %>%
  arrange(date) %>%
  mutate(
    ## Mark Feb 2020 as pre-covid anchor
    since_feb20 = case_when(
      date == pre_covid_ref ~ PAYEMS,
      date != pre_covid_ref ~ NA_real_
    ),
    ## Mark most recent month as current anchor
    since_now = case_when(
      date == max(date) ~ PAYEMS,
      date != max(date) ~ NA_real_
    )
  ) -> dat


## --- Create empty future rows for projection ---
month_dates <- seq(
  last(dat$date, 1) %m+% months(1),
  last(dat$date, 1) %m+% months(48),
  by = "month"
)

data.frame(
  date        = month_dates,
  PAYEMS      = rep(NA, length(month_dates)),
  since_feb20 = rep(NA, length(month_dates)),
  since_now   = rep(NA, length(month_dates))
) -> future

dat %>%
  rbind(future) -> dat_fut


## --- Compute projection growth rates ---
row_start_feb20 <- which(grepl(pre_covid_ref, dat_fut$date))
row_start_now   <- which(dat_fut$date == max(dat$date))

## Pre-covid 10-year average monthly job growth
dat_chg[which(dat_chg$date == pre_covid_ref):
        (which(dat_chg$date == pre_covid_ref) + 120), ] %>%
  select(date, nfp_chg_nsa, nfp_chg_sa) %>%
  summarize(avg = mean(nfp_chg_sa)) %>%
  unlist() -> pre_covid_10y_avg

## Last 3 months average job growth
dat_chg[1:3, ] %>%
  select(nfp_chg_sa) %>%
  summarize(avg_sa = mean(nfp_chg_sa)) %>%
  unlist() -> last_mths_avg


## --- Project forward from both anchor points ---
dat_fut %>%
  mutate(
    since_feb20 = case_when(
      date < pre_covid_ref  ~ PAYEMS,
      date >= pre_covid_ref ~
        pull(dat[row_start_feb20, "since_feb20"]) +
        (pre_covid_10y_avg * (row_number() - row_start_feb20))
    ),
    since_now = case_when(
      date < max(dat$date) ~ PAYEMS,
      date >= max(dat$date) ~
        pull(dat[row_start_now, "since_now"]) +
        (last_mths_avg * (row_number() - row_start_now))
    )
  ) -> df_final


## --- Chart labels ---
ctitle_1   <- "Nonfarm payrolls w/ current trajectory extrapolated"
unit_label <- "Payrolls, thousands, seasonally adjusted"


## --- Reshape for plotting ---
df_final %>%
  pivot_longer(-date, names_to = "key", values_to = "values") %>%
  mutate(
    key = case_when(
      key == "PAYEMS"      ~ "Actual total jobs",
      key == "since_feb20" ~ "Pre-pandemic trend",
      key == "since_now"   ~ "Last three months trend"
    )
  ) -> df_long

df_long$key <- factor(
  df_long$key,
  levels = c(
    "Pre-pandemic trend",
    "Last three months trend",
    "Actual total jobs"
  )
)


## --- Build animated trajectory chart ---
df_long %>%
  filter(
    date > (pre_covid_ref %m-% months(14)) &
    date < (current_month %m+% months(36))
  ) %>%
  ggplot(aes(date, values, color = key, linetype = key, size = key)) +
  geom_line() +
  geom_point(size = 2.5) +
  scale_linetype_manual(values = c("dashed", "dashed", "solid")) +
  scale_color_manual(values = c(
    cpal("green"),
    cpal("orange"),
    cpal("blue")
  )) +
  scale_size_manual(values = c(1, 1, 1.2)) +
  scale_x_date(
    expand      = c(0, 0),
    limits      = c(
      pre_covid_ref %m-% months(12),
      current_month %m+% months(24)
    ),
    date_labels = "'%y",
    date_breaks = "1 year"
  ) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title    = ctitle_1,
    subtitle = unit_label,
    caption  = source_1
  ) +
  my_theme +
  guides(
    color    = guide_legend(ncol = 1),
    linetype = guide_legend(ncol = 1)
  ) +
  theme(
    legend.key.height = unit(0.7, "cm"),
    plot.margin = unit(c(0.3, 0.5, 0.3, 0.5), "cm")
  ) +
  transition_reveal(date, keep_last = TRUE) -> chart1

anim1 <- animate(
  chart1,
  renderer  = gifski_renderer(loop = TRUE),
  nframe    = 260,
  width     = 615,
  height    = 460,
  units     = "px",
  res       = 72,
  type      = "cairo",
  end_pause = 200
)

anim1
```


An aggregate measure of labor market conditions, the Blanchflower-Levin employment gap, shows that tightness in the labor market has dissipated. Labor market "tightness," or conversely "slack," represents how close job market conditions are to what would be expected based on demographics, without either a glut of job opening or unemployed.


```{r blanchflower, echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE}
library(ggforce)
library(lubridate)


## --- Pull labor market data from FRED ---
## UNRATE = unemployment rate, CIVPART = LFPR,
## LNS12032194 = part-time for economic reasons, CLF16OV = civilian labor force

raw_dat <- getSymbols(
  c("UNRATE", "CIVPART", "LNS12032194", "CLF16OV"),
  src         = "FRED",
  auto.assign = TRUE
)

dat <- merge(UNRATE, CIVPART, LNS12032194, CLF16OV) %>%
  as_tibble(rownames = NA) %>%
  rownames_to_column() %>%
  rename(
    date = rowname,
    UR   = UNRATE,
    LFPR = CIVPART,
    PTER = LNS12032194,
    LF   = CLF16OV
  ) %>%
  mutate(
    date = as.Date(date, "%Y-%m-%d"),
    date = ceiling_date(date, "month") - 1
  ) %>%
  fill(UR, LFPR, PTER, LF, .direction = "down")


## --- Pull NAIRU (quarterly) and interpolate to monthly ---

raw_nairu <- getSymbols("NROU", src = "FRED", auto.assign = TRUE)

dat_nairu <- NROU %>%
  as_tibble(rownames = NA) %>%
  rownames_to_column() %>%
  rename(
    date  = rowname,
    NAIRU = NROU
  ) %>%
  mutate(date = as.Date(date, "%Y-%m-%d")) %>%
  filter(date < Sys.Date())

## Spline interpolation from quarterly to monthly
mthly_nairu <- seq(
  dat_nairu$date[1] + 1,
  tail(dat_nairu$date, 1) + 1,
  by = "1 month"
) - 1

nairu_m <- data.frame(
  date  = ceiling_date(mthly_nairu, "month") - 1,
  nairu = spline(
    dat_nairu,
    method = "fmm",
    xout   = mthly_nairu
  )$y
) %>%
  as_data_frame()


## --- Pull CBO potential LFPR (quarterly) and interpolate to monthly ---

cbo_dat <- read_csv("C:\\Users\\recuyell\\src\\data\\cbo_dat.csv") %>%
  select(date, potential_lfpr) %>%
  mutate(
    date = as.yearqtr(date, "%Yq%q"),
    date = as.Date(date, 1)
  )

monthly <- seq(
  cbo_dat$date[1] + 1,
  tail(cbo_dat$date, 1) + 1,
  by = "1 month"
) - 1

cbo_m <- data.frame(
  date     = monthly,
  cbo_lfpr = spline(
    cbo_dat,
    method = "fmm",
    xout   = monthly
  )$y
) %>%
  as_data_frame() %>%
  filter(date < Sys.Date())


## --- Merge all data and compute gap components ---

dat_tot <- dat %>%
  left_join(nairu_m, by = "date") %>%
  fill(nairu) %>%
  left_join(cbo_m, by = "date") %>%
  filter(
    date > "1979-12-31",
    date <= tail(dat$date, 1)
  ) %>%
  mutate(
    ## Adjust PTER for avg hours difference (part-time vs full-time)
    PTER_adj  = PTER * (5.48 / 8.08),
    PTER_perc = PTER_adj / LF
  )

## 1994-2007 average PTER share (pre-crisis benchmark)
pter_avg <- dat_tot %>%
  select(date, PTER_perc) %>%
  filter(
    date > "1993-12-31",
    date <= "2007-12-31"
  ) %>%
  summarize(avg = mean(PTER_perc)) %>%
  unlist()


## --- Compute Blanchflower-Levin gap components ---
## Long format for stacked bar chart
BL <- dat_tot %>%
  mutate(
    PTER_gap = (PTER_perc - pter_avg) * 100,
    UR_gap   = UR - nairu,
    LFPR_gap = cbo_lfpr - LFPR,
    total    = PTER_gap + UR_gap + LFPR_gap
  ) %>%
  select(date, PTER_gap, UR_gap, LFPR_gap) %>%
  rename(
    "Underemployment gap" = PTER_gap,
    "Unemployment gap"    = UR_gap,
    "Participation gap"   = LFPR_gap
  ) %>%
  pivot_longer(
    cols      = ends_with("gap"),
    names_to  = "key",
    values_to = "value"
  )

## Wide format with total line for dashed overlay
dat_line <- dat_tot %>%
  mutate(
    PTER_gap = (PTER_perc - pter_avg) * 100,
    UR_gap   = UR - nairu,
    LFPR_gap = cbo_lfpr - LFPR,
    total    = PTER_gap + UR_gap + LFPR_gap
  )


## --- Chart labels ---
ctitle_1   <- "Blanchflower-Levin employment gap"
unit_label <- "%"
source_1   <- "Source: BLS, FRED, @benbakkum."

start_date <- as.Date("1990-01-01")
end_date   <- tail(dat$date, 1)


## --- Build chart ---
BL_chart <- BL %>%
  filter(date >= start_date) %>% 
  mutate(placeholder = "X") %>%
  ggplot() +

  ## Stacked bar: gap components
  geom_bar(
    aes(fill = key, y = value, x = date),
    position = "stack",
    stat     = "identity",
    width    = 32
  ) +

  ## Inset zoom panel for recent years
  facet_zoom(
    xlim      = c(as.Date("2018/11/01"), Sys.Date()),
    ylim      = c(-2, 6),
    zoom.size = 0.4
  ) +

  ## Dashed total gap line
  geom_line(
    data     = dat_line %>% filter(date >= start_date),
    aes(x = date, y = total),
    size     = 0.8,
    alpha    = 0.5,
    linetype = "solid"
  ) +

  ## Scales
  scale_fill_manual(values = c("#33a02c", "#a6cee3", "#1f78b4")) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_date(
    date_labels = "'%y",
    expand      = expansion(mult = c(0, 0.03))
  ) +

  ## Theme — add horizontal spacing between legend items
  my_theme +
  theme(
    legend.text = element_text(
      size   = 10.4,
      color  = cpal("text"),
      family = "Open Sans",
      margin = margin(r = 50, unit = "pt")
    )
  ) +

  labs(
    title    = ctitle_1,
    subtitle = unit_label,
    caption  = paste0(
      "Data as of ",
      format(tail(dat_line$date, 1), "%b, %Y"),
      ". ",
      source_1
    )
  ) +
  transition_states(placeholder)

anim <- animate(
  BL_chart,
  renderer = gifski_renderer(loop = TRUE),
  fps      = 1,
  duration = 1,
  width    = 615,
  height   = 420,
  units    = "px",
  res      = 72,
  type     = "cairo"
)

anim
```

The various measures of tightness/slack above include:

- **Unemployment gap**: the difference between the unemployment rate and an estimate of the non-accelerating inflation rate of unemployment (NAIRU). NAIRU is a rough estimate of the level of unemployment that neither places upward or downward pressure on inflation.
- **Participation gap**: the difference between the labor force participation rate (LFPR) and the Congressional Budget Office’s estimate of the potential LFPR. The CBO’s "potential" version is their best guess at what that number should be given demographics (aging population, school enrollment etc). When actual participation falls below potential, it suggests there are people on the sidelines who would normally be in the workforce.
- **Underemployment gap**: the difference between the number of employees working part-time for economic reasons as a percentage of the labor force, adjusted for the difference in average hours worked by part-time and full-time employees, and the 1994-2007 average of this calculation.


The chart below shows the Beveridge curve, plotting the job openings rate (y-axis) against the unemployment rate (x-axis). It illustrates the inverse relationship between the two. When there are many openings as a percentage of the labor force, unemployment tends to be low, and vice versa. Recent data sits on a kink in the curve, suggesting that were the openings rate to continue to fall, the unemployment rate may increase at a faster pace.


```{r beveridge, echo=FALSE, warning=FALSE, message = FALSE, tidy=FALSE}

## --- Pull data ---------------------------------------------------------------

raw_bev <- getSymbols(c("UNRATE", "JTSJOR"), src = "FRED", auto.assign = TRUE)

dat_bev <- merge(UNRATE, JTSJOR) %>%
  as_tibble(rownames = NA) %>%
  rownames_to_column() %>%
  rename(
    date = rowname,
    ur   = UNRATE,
    jor  = JTSJOR
  ) %>%
  mutate(date = as.Date(date, "%Y-%m-%d")) %>%
  filter(!is.na(ur)) %>%
  ## Forward-fill openings rate so latest UR data uses most recent JOR
  fill(jor, .direction = "down") %>%
  filter(!is.na(jor)) %>%
  ## Remove covid outliers
  filter(date < as.Date("2020-03-01") | date > as.Date("2022-02-01")) %>%
  mutate(
    decade = paste0(floor(year(date) / 10) * 10, "s"),
    is_last     = row_number() == n(),
    placeholder = 1
  )


## --- Chart labels ------------------------------------------------------------

ctitle_1 <- "Beveridge curve"
unit_label <- "Job openings rate vs. unemployment rate"

last_pt  <- dat_bev %>% filter(is_last)
label_dt <- format(last_pt$date, "%b '%y")

decade_pal <- c(cpal("blue"), cpal("green"), cpal("red"), cpal("purple"),
                cpal("orange"), cpal("brown"))
decade_lvls <- sort(unique(dat_bev$decade))
decade_colors <- setNames(decade_pal[seq_along(decade_lvls)], decade_lvls)


## --- Build chart -------------------------------------------------------------

bev_chart <- dat_bev %>%
  filter(!is_last) %>%
  ggplot(aes(x = ur, y = jor)) +

  geom_point(
    aes(color = decade),
    size  = 2.5,
    alpha = 0.8
  ) +

  ## Latest data point (purple diamond, larger)
  geom_point(
    data  = last_pt,
    color = cpal("purple"),
    shape = 18,
    size  = 8
  ) +

  ## Arrow annotation to most recent point
  geom_curve(
    data = last_pt,
    aes(
      x    = ur + 1.4,
      y    = jor + 0.8,
      xend = ur + 0.15,
      yend = jor + 0.08
    ),
    curvature = -0.3,
    arrow     = arrow(length = unit(0.08, "inches"), type = "closed"),
    color     = cpal("purple"),
    size      = 0.4,
    inherit.aes = FALSE
  ) +
  geom_text(
    data = last_pt,
    aes(
      x     = ur + 1.45,
      y     = jor + 0.9,
      label = label_dt
    ),
    size     = 4,
    color    = cpal("purple"),
    fontface = "bold",
    family   = "Open Sans",
    inherit.aes = FALSE
  ) +

  ## Covid exclusion note
  annotate(
    "text",
    x      = Inf,
    y      = Inf,
    label  = "Mar '20 – Feb '22 excluded",
    hjust  = 1.05,
    vjust  = 1.5,
    size   = 3,
    color  = "grey50",
    family = "Open Sans"
  ) +

  ## Scales
  scale_color_manual(values = decade_colors) +
  scale_x_continuous(
    labels = function(x) paste0(x, "%"),
    breaks = function(x) seq(floor(x[1]), ceiling(x[2]), by = 1),
    expand = expansion(mult = 0.05)
  ) +
  scale_y_continuous(
    labels = function(x) paste0(x, "%"),
    breaks = function(x) seq(floor(x[1]), ceiling(x[2]), by = 1),
    expand = expansion(mult = 0.05)
  ) +

  ## Labels
  my_theme +
  theme(
    axis.title.x = element_text(
      size   = 10.8,
      color  = cpal("text"),
      family = "Open Sans"
    ),
    axis.title.y = element_text(
      size   = 10.8,
      color  = cpal("text"),
      family = "Open Sans"
    ),
    legend.margin = margin(t = 0, unit = "cm")
  ) +
  labs(
    title    = ctitle_1,
    subtitle = unit_label,
    x        = "Unemployment rate",
    y        = "Job openings rate",
    caption  = "Source: BLS, @benbakkum."
  ) +
  transition_states(placeholder)

anim <- animate(
  bev_chart,
  renderer = gifski_renderer(loop = TRUE),
  fps      = 1,
  duration = 1,
  width    = 615,
  height   = 420,
  units    = "px",
  res      = 72,
  type     = "cairo"
)

anim
```




